---
title: "tempssm: State-space modeling for temperature time series in R"
subtitle: "Version `r packageVersion('tempssm')`"
author: "Akira S. Hirao, Shinya Baba, and Momoko Ichinokawa"
date: "`r Sys.Date()`"
output:
#  rmarkdown::html_vignette: default
#  pdf_document:
#    latex_engine: lualatex
#    ##toc: true
  md_document:
    variant: gfm+tex_math_dollars
vignette: >
  %\VignetteIndexEntry{tempssm: State-space modeling for temperature time series in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Summary

`tempssm` is an R package for state-space modeling of environmental
temperature time series, including air and water temperature observations. It
provides a practical framework for assessing how long-term trend, seasonal
variation, autoregressive dependence, and optional exogenous effects contribute
to observed temporal variation. The package facilitates the application of
linear Gaussian state-space models estimated by Kalman filtering and smoothing,
using the `KFAS` package as the computational backend (Helske, 2017).

# Key Features

- Fits linear Gaussian state-space models to environmental temperature time
  series.
- Represents temperature dynamics using interpretable latent components:
  long-term trend, seasonal variation, autoregressive dependence, and optional
  exogenous effects.
- Supports arbitrary seasonal frequencies, while the current examples
  and validation focus primarily on monthly temperature data.
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1)).
- Includes time-series cross-validation tools for model evaluation.

# Input Data Format

R `ts` objects are the primary input format for `tempssm`. The temperature
series passed to `tempssm()` should be supplied as a univariate `ts` object,
and optional exogenous variables can be supplied as univariate or multivariate
`ts` objects. The `ts` class is base R's standard format for regularly spaced
time series (see the `stats::ts` documentation:
<https://search.r-project.org/R/refmans/stats/html/ts.html>). The package also
provides utility functions for converting common tabular and observational
data formats into `ts` objects before model fitting.

The seasonal cycle used by `tempssm()` is taken from the `frequency` attribute
of the input `ts` object. For example, `frequency = 12` represents monthly
data.

# Prior Art and Scope

`tempssm` provides a domain-focused workflow for analyzing temperature time
series with linear Gaussian state-space models. It brings together model
construction, component extraction, uncertainty summaries, residual
diagnostics, visualization, and time-series cross-validation in a single R
package interface tailored to temperature applications.

The package builds on established statistical methodology, including linear
Gaussian state-space modeling, Kalman filtering, and Kalman smoothing. Model
estimation is handled through the `KFAS` package, which provides a general
framework for state-space models in R.

The initial implementation was adapted from the supplementary code provided by
Baba (2024), accompanying Baba et al. (2024), which analyzed sea temperature
trends using a linear Gaussian state-space model. The supplementary code is
publicly available at:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

Compared with that prior implementation, `tempssm` extends the workflow into a
reusable R package interface with input validation, documented S3 methods,
tests, diagnostics, cross-validation utilities, and examples for broader
temperature time-series analysis.

# Model Overview

The model implemented in `tempssm` can be viewed as an extension of the Basic
Structural Time Series Model (BSTSM), a standard state-space formulation that
represents an observed time series using latent trend, seasonal, and irregular
components. Following the temperature time-series application of Baba et al.
(2024), `tempssm` keeps this interpretable decomposition and adds
autoregressive dependence and optional exogenous effects. This makes it useful
for separating long-term temperature change, seasonal cycles, and short-term
departures.

The model is estimated with `KFAS`, and `tempssm()` returns both filtering and
smoothing estimates. Unless otherwise stated, the summaries, diagnostics, and
plots in this vignette use smoothed state estimates.

The full mathematical specification, including the observation equation, state
decomposition, seasonal constraint, autoregressive component, exogenous
component, parameter count, and estimation procedure, is described in the
separate model specification document, available as a PDF in the package
repository:

<https://github.com/akihirao/tempssm/blob/main/vignettes/model-specification.pdf>

# Setup

Load `tempssm` before running the examples below.

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
```

The objective of this practice is to demonstrate the basic application
of a linear Gaussian state-space model to a univariate temperature time
series. This example introduces the basic workflow for fitting a `tempssm`
model, inspecting the summary output, plotting latent components, checking
residual diagnostics, and making short-term predictions.

# Example Dataset

This tutorial uses `temp_MtFuji`, a long-term monthly air temperature time
series observed at the summit of Mt. Fuji, Japan.

- **Dataset**: Monthly mean air temperature at the summit of Mt. Fuji, Japan
- **Format**: Univariate `ts` object
- **Frequency**: 12 (monthly)
- **Unit**: Degrees Celsius
- **Period**: July 1932 to June 2026

The dataset was created by processing content from the Japan Meteorological
Agency (JMA), Past Weather Data Download page:
<https://www.data.jma.go.jp/risk/obsdl/index.php>. The series may contain
missing values from the source data; values with JMA quality flags below 8,
where 8 indicates a normal value with no quality problem, are also treated as
missing values.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(temp_MtFuji) # load a ts object of temperature at Mt. Fuji
head(temp_MtFuji)
summary(temp_MtFuji)
```

This dataset contains missing observations. `tempssm()` can retain missing
values in the response temperature series and treat them as unobserved
responses during Kalman filtering and smoothing.

# Plot the Time Series

We begin by visualizing the monthly air temperature time series to examine its
overall structure, including apparent trends, seasonal variability, and whether
missing observations are present.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_mtfuji_temp <- forecast::autoplot(temp_MtFuji) +
  ggplot2::labs(
    y = expression(Temperature ~ (degree * C)),
    x = "Time (year)"
  ) +
  ggplot2::ggtitle("Monthly air temperature at the summit of Mt. Fuji") +
  ggplot2::theme_classic()

plot(plt_mtfuji_temp)
```

```{r echo = FALSE}
mtfuji_temp_mean <- mean(temp_MtFuji, na.rm = TRUE)
mtfuji_temp_missing <- sum(is.na(temp_MtFuji))
```

The overall mean air temperature is approximately
`r sprintf("%.1f", mtfuji_temp_mean)` °C, and a clear annual seasonal pattern
is visible. The series contains `r mtfuji_temp_missing` missing observations.
Although the observed temperature appears to increase gradually over the long
term, year-to-year variation and strong seasonality make 
the underlying trend difficult to assess from the raw time series alone.

# Fit a State-Space Model

When a `ts` object containing temperature time-series data (here,
`temp_MtFuji`) is passed to the core function `tempssm()`, model construction
and parameter estimation are performed together. The returned S3 object of
class `tempssm`
(here, `res`) stores the filtering and smoothing estimates, as well as the
constructed model and input data. By default, `tempssm()` fits a first-order
autoregressive model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# model with first-order autoregressive component
res <- tempssm(temp_MtFuji) # AR(1), the default model
summary(res)
```

First, confirm from the summary output that the model has converged
(`Converged: TRUE`). The summary also reports the log-likelihood, parameter
count (`k`), likelihood type, and number of diffuse initial states. The
estimated parameters include the observation error variance (`H`) and the
process error variances for the long-term trend (`Q trend`), seasonal
component (`Q season`), and autoregressive component (`Q ar`), as well as the
autoregressive coefficient (`AR1` in the default model). See the detailed
manual for extracting and using these quantities directly; its location is
listed at the end of this quick tutorial.

# Plot Model Components

We visualize the estimated long-term evolution of temperature levels
and their rates of change (drift) by extracting the corresponding latent
components from the state-space model. Additionally, seasonal variability and
autoregressive dependence are plotted out, allowing the underlying trend
behavior to be examined more clearly.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot all components at once
plot(res)
```

The level component summarizes gradual changes in the underlying temperature
level after removing the dominant seasonal pattern. In this example, the
estimated long-term level indicates a gradual increase in temperature 
over the study period. The drift component is positive on average but small
relative to the seasonal variation. The seasonal component captures the
recurring annual cycle, while the autoregressive component represents
shorter-term departures from the trend and seasonal pattern. The shaded gray
areas represent 95% confidence intervals for the estimated latent states,
illustrating the uncertainty associated with each estimated component.

The standard plotting interface is `plot(res)`. The explicit helper
`plot_tempssm_components(res)` produces the same component plot and can be
useful in scripts where a descriptive function name is preferred. The
ggplot2-style S3 interface `ggplot2::autoplot(res)` is also available. These
interfaces return a faceted `ggplot` object, so selected components can be
stored and customized with standard ggplot2 layers, for example,
`plot_tempssm_components(res, component = c("level", "drift")) +
ggplot2::theme_bw()`.


# Check Residual Diagnostics

The package provides diagnostic tools for checking whether the fitted model 
has left notable structure in the residuals. In particular, residual 
time-series plots, residual autocorrelation, residual distributions, and 
Ljung-Box tests can be used to assess remaining temporal dependence and 
departures from the Gaussian error assumption.

```{r message = FALSE, warning = FALSE, echo = TRUE}
r <- get_tempssm_residuals(res)
plot_tempssm_residuals(r, frequency = frequency(res$temp_data))
diag <- diagnose_residuals(res)
diag
```

Here, `get_tempssm_residuals()` extracts standardized recursive residuals.
`plot_tempssm_residuals()` displays the residual time series in the upper
panel, the ACF plot in the lower-left panel, and the residual frequency
distribution in the lower-right panel. `diagnose_residuals()` returns a compact
table with the Ljung-Box test and residual kurtosis. For monthly data, the
default Ljung-Box lag is 12. These plots and the test result should be checked
for any notable residual patterns. In this example, the Ljung-Box test
indicated no significant residual autocorrelation up to lag 12 (P > 0.05).

For more detailed discussion of residual diagnostics, see the detailed manual.

# Extract Components

The fitted object stores the smoothed latent states in `res$kfs$alphahat`.
Each column corresponds to one state component in the state-space model, such
as the level, drift, seasonal states, and autoregressive state. Looking at the
first few rows is useful for understanding how the model output is organized.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res$kfs$alphahat
head(alpha_hat)
```

For routine use, helper functions provide a simpler way to extract individual
components as `ts` objects with the original time index. The level component
represents the estimated long-term temperature level, while the drift
component represents its rate of change per year.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimate of level component
level_ts <- get_level_ts(res)

# Smoothing estimate of drift component
drift_ts <- get_drift_ts(res)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts)
mean_drift_year
```
 
Average annual change in air temperature is approximately
`r sprintf("%.2f", mean_drift_year)` °C. This value is the mean of the
estimated drift component over the full observation period, and can be read as
a model-based summary of the average long-term rate of temperature change.

# Make Short-Term Predictions

A fitted `tempssm` object can also be passed to `predict()` to obtain
short-term predictions. By default, `predict(res)` returns a one-step-ahead
prediction beyond the end of the observed series. This is useful for visual
checks of how the fitted model extrapolates the estimated level, seasonal, and
autoregressive components.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_1 <- predict(res)
pred_1
```

Predictions for multiple future time points can be requested by setting the
`n.ahead` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12 <- predict(res, n.ahead = 12)
pred_12
```

These predictions should be interpreted as model-based extrapolations rather
than definitive forecasts. Uncertainty generally increases as the prediction
horizon becomes longer, and long-horizon predictions can be sensitive to model
assumptions about trend, seasonality, and autoregressive dependence.

# Read Your Own CSV Data

For monthly temperature data stored in a CSV file, prepare columns named
`Year`, `Month`, and `Temp`. Use `NA` for missing temperature values, and keep
the corresponding `Year` and `Month` entries.

```text
Year,Month,Temp
2010,8,13.6
2010,9,6.8
2010,10,NA
2010,11,-1.4
...
```

The CSV file should be comma-separated and UTF-8 encoded. The package includes
an example CSV file in `inst/extdata`. The helper function
`read_monthly_temp_ts()` reads this type of CSV file and converts it into an R
`ts` object for use with `tempssm()`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file(
  "extdata",
  "example_monthly_temp.csv",
  package = "tempssm"
)

csv_temp <- tempssm::read_monthly_temp_ts(path)
head(csv_temp)
```

# Further Reading

The examples above illustrate the basic univariate workflow for fitting,
diagnosing, visualizing, and making short-term predictions from a `tempssm`
model. The detailed manual extends this workflow to exogenous variables,
additional model diagnostics, and time-series cross-validation:

https://github.com/akihirao/tempssm/blob/main/tools/manual/tempssm_manual.pdf
