---
title: "tempssm: State-space modeling for temperature time series in R"
subtitle: "Version `r packageVersion('tempssm')`"
author: "Akira S. Hirao, Shinya Baba, and Momoko Ichinokawa"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
    ##toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm+tex_math_dollars
vignette: >
  %\VignetteIndexEntry{tempssm: State-space modeling for temperature time series in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Summary

R package **tempssm** provides tools for state-space analysis of temperature
time series, focusing on linear Gaussian state-space models estimated via
Kalman filtering and smoothing as implemented in the KFAS package
(Helske, 2017).

### Key features

- Designed for temperature time series with arbitrary seasonal frequencies;
  currently validated primarily on monthly data
- Estimates latent states using linear Gaussian state-space models
  combined with Kalman filtering and smoothing
- Models temperature dynamics as a sum of interpretable latent components,
  including a long-term trend, seasonal cycle, autoregressive structure,
  and optional exogenous effects
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1))
- Implements time-series cross-validation for model evaluation

# Prior Art and Scope

`tempssm` provides a domain-focused workflow for analyzing temperature time
series with linear Gaussian state-space models. It brings together model
construction, component extraction, uncertainty summaries, residual
diagnostics, visualization, and time-series cross-validation in a single R
package interface tailored to temperature applications.

The package builds on established statistical methodology, including linear
Gaussian state-space modeling, Kalman filtering, and Kalman smoothing. Model
estimation is handled through the `KFAS` package, which provides a general
framework for state-space models in R.

The initial implementation was adapted from the supplementary code provided by
Baba (2024), accompanying Baba et al. (2024), which analyzed sea temperature
trends using a linear Gaussian state-space model. The supplementary code is
publicly available at:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

Compared with that prior implementation, `tempssm` extends the workflow into a
reusable R package interface with input validation, documented S3 methods,
tests, diagnostics, cross-validation utilities, and examples for broader
temperature time-series analysis.

The next section describes the state-space model used by `tempssm` and
clarifies how the trend, seasonal, autoregressive, and exogenous components are
represented.

# State-Space Model in **tempssm**

We consider an extended version of the Basic Structural Time Series Model
(BSTSM) to describe temperature time series with explicit long-term trend,
seasonal variability, and auto-regressive component. The observation equation
is given by
$$
y_t = \alpha_t + v_t. \tag{1}
$$
where $t$ denotes the time index, $y_t$ is the observed temperature,
$\alpha_t$ is the latent state, and $v_t$ is the observation error.

The latent state is decomposed as 
$$
\alpha_t = \mu_t + s_t + r_t + E_t. \tag{2}
$$
where $\mu_t$ is the level (trend), $s_t$ the seasonal component, $r_t$
a stationary autoregressive component, and $E_t$ the contribution of
exogenous variables.


The level component follows a second-order stochastic process:
$$
\mu_t = 2\mu_{t-1} - \mu_{t-2} + \zeta_t,
\qquad \zeta_t \sim \mathcal{N}(0, \sigma_\zeta^2). \tag{3}
$$
where $\zeta_t$ is the process error of the level component. 

The seasonal component is modeled with a sum-to-zero constraint:
$$
s_t = - \sum_{i=t-f}^{t-1} s_i + \omega_t,
\qquad \omega_t \sim \mathcal{N}(0, \sigma_\omega^2). \tag{4}
$$
where $\omega_t$ is the process error of the seasonal component and $f$
denotes the seasonal frequency (e.g., $f=12$ for monthly data). This
formulation ensures that seasonal effects are identifiable and comparable
across different temporal resolutions. By enforcing the sum-to-zero constraint
over one full seasonal cycle, the seasonal component captures recurring
deviations from the underlying trend without introducing long-term drift. In
models without a seasonal component, the seasonal term $s_t$ is set to zero
and omitted from the satate equation.

The autoregressive component follows an $l$th-order autoregressive (AR) process:
$$
r_t = \phi_1 r_{t-1} + \phi_2 r_{t-2} + \cdots + \phi_l r_{t-l} + \tau_t,
\qquad \tau_t \sim \mathcal{N}(0, \sigma_\tau^2). \tag{5}
$$
Here, $\tau_t$ denotes the process error of the autoregressive component. The
process error terms ($\zeta_t$, $\omega_t$, and $\tau_t$) are assumed to be
mutually independent and normally distributed.

The exogenous component is defined as
$$
E_t = \beta_1 x_{1,t} + \beta_2 x_{2,t} + \cdots + \beta_m x_{m,t}. \tag{6}
$$
The exogenous term $E_t$ represents the influence of external factors, which
may include large-scale climate indices, regional environmental variables, or
other physically motivated predictors relevant to the observed temperature time
series. Each exogenous variable enters the model linearly through a
time-invariant regression coefficient ($\beta_1, \beta_2, \ldots, \beta_m$),
allowing both the magnitude and direction of its contribution to be estimated
jointly with the latent state components.

When applying the model to observational data, the total number of parameters
to be estimated ($k$) depends on the order of autoregressive component and the
number of exogenous variables included. For example, in a model with a
second-order autoregressive component and no exogenous covariates, six
parameters are estimated: the observation error variance, the process error
variances associated with the level, seasonal, and autoregressive components,
and the first- and second-order autoregressive coefficients ($\phi_1$ and
$\phi_2$). In the general case with an $l$-order autoregressive component and
$m$ exogenous variables, a total number of parameters is $k = 4 + l + m$.

The core implementation of parameter estimation procedure is based on the
supplementary code provided by Baba et al. (2024). Parameter estimation is
performed using a two-step optimization strategy recommended by Helske (2017).
In the first step, model parameters are estimated using the Nelder-Mead method
(Nelder & Mead, 1965) with user-specified initial values. The resulting
estimates are then used as initial values in a second optimization step based
on the BFGS algorithm (Shanno, 1970). The main model-fitting function,
`tempssm()`, returns both filtering and smoothing estimates. Unless otherwise
stated, all results presented in this vignette are based on the smoothed
estimates.


# How to use

## Set Ennvironment

Load the following libraries for executing 'How to use'.

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
```

## Input Data Format

Input data for **tempssm** must be supplied as an R `ts` object,
which represents a regularly spaced time series
(see `?stats::ts` or
<https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html>).

To support data preparation, the package includes utility functions
that convert external observational data into `ts` objects suitable
for model fitting (see Appendix).


## Practice: Applying State-Space Model to a Univariate Temperature Time Series

### Objective    

The objective of this practice is to demonstrate the basic application
of a linear Gaussian state-space model to a univariate temperature time
series. This example serves as an introduction to the modeling framework
and highlights the role of autoregressive dynamics without the inclusion
of exogenous variables.

### Loading the Dataset    

A sample sea surface temperature (SST) dataset is included in the package.

- **Dataset**: Monthly sea surface temperature (SST) off Niigata, Japan  
- **Unit**: Degrees Celsius  
- **Period**: February 2002 to December 2023  

This dataset is derived from observations archived at Japan Oceanographic Data
Center (JODC), Hydrographic and Oceanographic Department, Japan Coast Guard.
Original daily SST data were obtained from
<https://www.jodc.go.jp/jodcweb/JDOSS/index.html> and aggregated into monthly
means.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(niigata_sst) # load a ts object of SST off Niigata
head(niigata_sst)
summary(niigata_sst)
```

The dataset includes two missing observations. Even if missing observations
was in your dataset included, there are retained and handled explicitly within
the state-space modeling framework.

### Plotting the Monthly SST Time Series    

We begin by visualizing the monthly SST time series to examine its
overall structure, including apparent trends, seasonal variability,
and missing observations.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_niigata_sst <- forecast::autoplot(niigata_sst) +
  ggplot2::labs(
    y = expression(Temperature ~ (degree * C)),
    x = "Time (year)"
  ) +
  ggplot2::ggtitle("Monthly SST off Niigata, Japan") +
  ggplot2::theme_classic()

plot(plt_niigata_sst)
```

```{r echo = FALSE}
niigata_sst_mean <- mean(niigata_sst, na.rm = TRUE)
```

The overall mean SST is approximately `r sprintf("%.0f", niigata_sst_mean)` °C,
and a clear seasonal pattern is visible. The series contains two missing
observations, in September 2005 and February 2006. Although SST appears to be
higher near the end of the series than near the beginning, year-to-year
variability is also evident, making the long-term trend difficult to assess
from the raw time series alone.

### Applying a Linear Gaussian State-Space Model

When a ts object containing temperature time-series data (here, `niigata_sst`) 
is passed to the core function tempssm(), model construction and parameter 
estimation are performed together. The returned S3 object of class tempssm 
(here, `res`) stores the filtering and smoothing estimates, as well as the
constructed model and input data. By default, tempssm() fits a first-order autoregressive model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# model with first-order autoregressive component
res <- tempssm(niigata_sst) # AR(1), the default model
summary(res)
```

From the summary output, confirm that the model has converged (Converged: 
TRUE). The output also reports statistics such as the number of parameters 
(k), the log-likelihood, the likelihood type, and the number of diffuse initial 
states. The parameter estimates include the observation error variance (H), 
the process error variance of the long-term trend component (Q trend), 
the process error variance of the seasonal component (Q season), the process 
error variance of the autoregressive component, and the first-order 
autoregressive coefficient (AR1).

The log-likelihood and the associated number of estimated parameters can also
be extracted directly from the fitted `tempssm` object using `logLik()`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
ll <- logLik(res)
ll
attr(ll, "df") # number of parameters
```

By default, `tempssm()` uses the KFAS marginal likelihood for parameter
estimation. The selected likelihood type is retained for `logLik()` and
`summary()`. The package intentionally does not compute AIC for `tempssm`
objects. The log-likelihood and parameter count remain available through
`logLik()` for users who need them for their own model-assessment workflows.
The diffuse likelihood remains available by fitting the model with
`marginal = FALSE`.

### Plotting Level, Drift, Seasonal, and Auto-Regressive Components    

We visualize the estimated long-term evolution of temperature levels
and their rates of change (drift) by extracting the corresponding latent
components from the state-space model. Additionally, seasonal variability and
autoregressive dependence are plotted out, allowing the underlying trend
behavior to be examined more clearly.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot all components at once
plot(res)
```

The level component shows a persistent upward trend in sea surface
temperature over the study period, while the drift component indicates
a relatively stable positive rate of change. The shaded gray areas
represent 95% confidence intervals for the estimated latent states,
illustrating the uncertainty associated with each of estimated components.

Although the observed time series contains missing values, the state-space 
framework allows latent states to be estimated for the entire time span, 
including unobserved periods.

The standard plotting interface is `plot(res)`. The ggplot2-style interface
`autoplot(res)` is also available and produces the same component plot by
default. It returns a faceted `ggplot` object, so selected components can be
stored and customized with standard ggplot2 layers, for example,
`autoplot(res, component = c("level", "drift")) + ggplot2::theme_bw()`.


### Simple Model Diagnostics    

The package provides diagnostic tools for checking whether the fitted model 
has left notable structure in the residuals. In particular, residual 
time-series plots, residual autocorrelation, residual distributions, and 
Ljung-Box tests can be used to assess remaining temporal dependence and 
departures from the Gaussian error assumption.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plot_tempssm_residual_diagnostics(res)
```

In the model diagnostic plot, the upper panel shows the residual time series, 
the lower-left panel shows the residual autocorrelation plot (ACF plot), and 
the lower-right panel shows the residual frequency distribution.
These plots should be checked for any notable residual patterns.

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag <- diagnose_residuals(res)
print(diag)
```

The `lb_stat`, `lb_lag`, and `lb_pvalue` columns correspond to the Ljung-Box
test statistic, the lag used in the test, and its P-value, respectively. For
monthly time series, `diagnose_residuals()` uses lag 12 by default. In this
example, the Ljung-Box test indicated no significant residual autocorrelation 
up to lag 12 (P > 0.05).

### Estimated Parameters and Latent-State Components    

The long-term trend component and its rate of change (drift) can be extracted 
as ts objects as follows.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res$kfs$alphahat
head(alpha_hat)

# 　Smoothing estimate of level component
level_ts <- get_level_ts(res)

# 　Smoothing estimate of drift component
drift_ts <- get_drift_ts(res)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts)
print(mean_drift_year)
```
 
Average annual increase in SST is approximately
`r sprintf("%.2f", mean_drift_year)` °C.

### Short-Term Prediction

A fitted `tempssm` object can also be passed to `predict()` to obtain
short-term predictions. By default, `predict(res)` returns a one-step-ahead
prediction beyond the end of the observed series. This is useful for visual
checks of how the fitted model extrapolates the estimated level, seasonal, and
autoregressive components.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_1 <- predict(res)
pred_1
```

Predictions for multiple future time points can be requested by setting the
`n.ahead` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12 <- predict(res, n.ahead = 12)
pred_12
```

These predictions should be interpreted as model-based extrapolations rather
than definitive forecasts. Uncertainty generally increases as the prediction
horizon becomes longer, and long-horizon predictions can be sensitive to model
assumptions about trend, seasonality, and autoregressive dependence.

The examples above illustrate the basic univariate workflow for fitting,
diagnosing, visualizing, and making short-term predictions from a `tempssm`
model. The detailed manual extends this workflow to exogenous variables,
additional model diagnostics, and time-series cross-validation:

https://github.com/akihirao/tempssm/blob/main/tools/manual/tempssm_manual.pdf
