---
title: "tempssm: State-space modeling for temperature time series in R"
subtitle: "Version `r packageVersion('tempssm')`"
author: "Akira S. Hirao, Shinya Baba, and Momoko Ichinokawa"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
#  pdf_document:
#    latex_engine: lualatex
#    ##toc: true
#  md_document:
#    variant: gfm+tex_math_dollars
vignette: >
  %\VignetteIndexEntry{tempssm: State-space modeling for temperature time series in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Summary

`tempssm` is an R package for state-space modeling of environmental
temperature time series, including air and water temperature observations. It
provides a practical framework for assessing how long-term trend, seasonal
variation, autoregressive dependence, and optional exogenous effects contribute
to observed temporal variation. The package facilitates the application of
linear Gaussian state-space models estimated by Kalman filtering and smoothing,
using the `KFAS` package as the computational backend (Helske, 2017).

# Key Features

- Fits linear Gaussian state-space models to environmental temperature time
  series.
- Represents temperature dynamics using interpretable latent components:
  long-term trend, seasonal variation, autoregressive dependence, and optional
  exogenous effects.
- Supports arbitrary seasonal frequencies, while the current examples
  and validation focus primarily on monthly temperature data.
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1)).
- Includes time-series cross-validation tools for model evaluation.

# Input Data Format

The core modeling functions in `tempssm` expect input time series to be
supplied as R `ts` objects. The `ts` class is base R's standard format for
regularly spaced time series (see the `stats::ts` documentation:
<https://search.r-project.org/R/refmans/stats/html/ts.html>). The package also
provides utility functions for converting common tabular and observational data
formats into `ts` objects before model fitting.

The seasonal cycle used by `tempssm()` is taken from the `frequency` attribute
of the input `ts` object. For example, `frequency = 12` represents monthly
data.

# Prior Art and Scope

`tempssm` provides a domain-focused workflow for analyzing temperature time
series with linear Gaussian state-space models. It brings together model
construction, component extraction, uncertainty summaries, residual
diagnostics, visualization, and time-series cross-validation in a single R
package interface tailored to temperature applications.

The package builds on established statistical methodology, including linear
Gaussian state-space modeling, Kalman filtering, and Kalman smoothing. Model
estimation is handled through the `KFAS` package, which provides a general
framework for state-space models in R.

The initial implementation was adapted from the supplementary code provided by
Baba (2024), accompanying Baba et al. (2024), which analyzed sea temperature
trends using a linear Gaussian state-space model. The supplementary code is
publicly available at:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

Compared with that prior implementation, `tempssm` extends the workflow into a
reusable R package interface with input validation, documented S3 methods,
tests, diagnostics, cross-validation utilities, and examples for broader
temperature time-series analysis.

The next section gives a brief model overview. A separate model specification
vignette describes the mathematical formulation in more detail.

# Model Overview

The model implemented in `tempssm` can be viewed as an extension of the Basic
Structural Time Series Model (BSTSM), a standard state-space formulation that
represents an observed time series using latent trend, seasonal, and irregular
components. Following the temperature time-series application of Baba et al.
(2024), `tempssm` keeps this interpretable decomposition and adds
autoregressive dependence and optional exogenous effects. This makes it useful
for separating long-term temperature change, seasonal cycles, and short-term
departures.

The model is estimated with `KFAS`, and `tempssm()` returns both filtering and
smoothing estimates. Unless otherwise stated, the summaries, diagnostics, and
plots in this vignette use smoothed state estimates.

The full mathematical specification, including the observation equation, state
decomposition, seasonal constraint, autoregressive component, exogenous
component, parameter count, and estimation procedure, is described in the
separate model specification document, available as a PDF in the package
repository:

<https://github.com/akihirao/tempssm/blob/main/vignettes/model-specification.pdf>

# Setup

Load `tempssm` before running the examples below.

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
```

The objective of this practice is to demonstrate the basic application
of a linear Gaussian state-space model to a univariate temperature time
series. This example introduces the basic workflow for fitting a `tempssm`
model, inspecting the summary output, plotting latent components, checking
residual diagnostics, and making short-term predictions.

# Example Dataset

A sample sea surface temperature (SST) dataset is included in the package.

- **Dataset**: Simulated monthly sea surface temperature (SST) off Jogashima,
  Miura City, Kanagawa Prefecture, Japan
- **Format**: Univariate `ts` object
- **Frequency**: 12 (monthly)
- **Unit**: Degrees Celsius
- **Period**: January 1998 to February 2023

This dataset is a simulated monthly SST time series based on the state-space
model analysis of sea temperature off Jogashima reported by Baba et al. (2024).
It is distributed with the supplementary materials and prototype code for the
motivating study:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(sst_jogashima) # load a ts object of SST off Jogashima
head(sst_jogashima)
summary(sst_jogashima)
```

This simulated dataset contains no missing observations. In general,
`tempssm()` can retain missing values in the response temperature series and
treat them as unobserved responses during Kalman filtering and smoothing.

# Plot the Time Series

We begin by visualizing the monthly SST time series to examine its
overall structure, including apparent trends, seasonal variability,
and whether missing observations are present.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_jogashima_sst <- forecast::autoplot(sst_jogashima) +
  ggplot2::labs(
    y = expression(Temperature ~ (degree * C)),
    x = "Time (year)"
  ) +
  ggplot2::ggtitle("Simulated monthly SST off Jogashima, Japan") +
  ggplot2::theme_classic()

plot(plt_jogashima_sst)
```

```{r echo = FALSE}
sst_jogashima_mean <- mean(sst_jogashima, na.rm = TRUE)
```

The overall mean SST is approximately `r sprintf("%.1f", sst_jogashima_mean)`
°C, and a clear seasonal pattern is visible. The series contains no missing
observations. SST appears to decline gradually from the beginning of the series
to around 2010 and then increase thereafter. However, year-to-year variability
is also evident, making it difficult to identify a clear long-term pattern from
the raw time series alone.

# Fit a State-Space Model

When a `ts` object containing temperature time-series data
(here, `sst_jogashima`) is passed to the core function `tempssm()`, model
construction and parameter estimation are performed together. The returned S3
object of class `tempssm`
(here, `res`) stores the filtering and smoothing estimates, as well as the
constructed model and input data. By default, `tempssm()` fits a first-order
autoregressive model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# model with first-order autoregressive component
res <- tempssm(sst_jogashima) # AR(1), the default model
summary(res)
```

First, confirm from the summary output that the model has converged
(`Converged: TRUE`). The summary also reports the log-likelihood, parameter
count (`k`), likelihood type, and number of diffuse initial states. The
estimated parameters include the observation error variance (`H`) and the
process error variances for the long-term trend (`Q trend`), seasonal
component (`Q season`), and autoregressive component (`Q ar`), as well as the
autoregressive coefficient (`AR1` in the default model). See the detailed
manual for extracting and using these quantities directly; its location is
listed at the end of this quick tutorial.

# Plot Model Components

We visualize the estimated long-term evolution of temperature levels
and their rates of change (drift) by extracting the corresponding latent
components from the state-space model. Additionally, seasonal variability and
autoregressive dependence are plotted out, allowing the underlying trend
behavior to be examined more clearly.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot all components at once
plot(res)
```

The level component suggests a long-term pattern that changes around 2008:
the underlying SST level decreases during the first part of the series and
then increases during the latter part. The drift component shows a
corresponding pattern, with mostly negative values before around 2008 and
positive values thereafter. The seasonal component captures the recurring
annual cycle, while the autoregressive component represents shorter-term
departures from the trend and seasonal pattern. The shaded gray areas
represent 95% confidence intervals for the estimated latent states,
illustrating the uncertainty associated with each estimated component.
Because this is a simulated dataset, these component plots should be
interpreted as an illustration of the model output rather than as direct
estimates from the original observational record.

The standard plotting interface is `plot(res)`. The explicit helper
`plot_tempssm_components(res)` produces the same component plot and can be
useful in scripts where a descriptive function name is preferred. The
ggplot2-style S3 interface `ggplot2::autoplot(res)` is also available. These
interfaces return a faceted `ggplot` object, so selected components can be
stored and customized with standard ggplot2 layers, for example,
`plot_tempssm_components(res, component = c("level", "drift")) +
ggplot2::theme_bw()`.


# Check Residual Diagnostics

The package provides diagnostic tools for checking whether the fitted model 
has left notable structure in the residuals. In particular, residual 
time-series plots, residual autocorrelation, residual distributions, and 
Ljung-Box tests can be used to assess remaining temporal dependence and 
departures from the Gaussian error assumption.

```{r message = FALSE, warning = FALSE, echo = TRUE}
r <- get_tempssm_residuals(res)
lb_lag <- frequency(res$temp_data)
forecast::checkresiduals(r, lag = lb_lag, test = "LB")
```

Here, `get_tempssm_residuals()` explicitly extracts the standardized recursive
residuals from the fitted model. `forecast::checkresiduals()` then displays
the residual time series, residual autocorrelation plot (ACF plot), and
residual frequency distribution, together with a Ljung-Box test. The lag is
set to the seasonal frequency of the input data; for monthly data, this uses
lag 12. These plots and the test result should be checked for any notable
residual patterns. In this example, the Ljung-Box test indicated no
significant residual autocorrelation up to lag 12 (P > 0.05).

For tabular residual diagnostic summaries, see `diagnose_residuals()` in the
detailed manual.

For repeated checks across many fitted models, the convenience wrapper
`plot_tempssm_residual_diagnostics(res)` provides the same type of residual
diagnostic plot.

# Extract Components

The long-term trend component and its rate of change (drift) can be extracted 
as ts objects as follows.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res$kfs$alphahat
head(alpha_hat)

# 　Smoothing estimate of level component
level_ts <- get_level_ts(res)

# 　Smoothing estimate of drift component
drift_ts <- get_drift_ts(res)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts)
mean_drift_year
```
 
Average annual increase in SST is approximately
`r sprintf("%.2f", mean_drift_year)` °C.

# Make Short-Term Predictions

A fitted `tempssm` object can also be passed to `predict()` to obtain
short-term predictions. By default, `predict(res)` returns a one-step-ahead
prediction beyond the end of the observed series. This is useful for visual
checks of how the fitted model extrapolates the estimated level, seasonal, and
autoregressive components.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_1 <- predict(res)
pred_1
```

Predictions for multiple future time points can be requested by setting the
`n.ahead` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12 <- predict(res, n.ahead = 12)
pred_12
```

These predictions should be interpreted as model-based extrapolations rather
than definitive forecasts. Uncertainty generally increases as the prediction
horizon becomes longer, and long-horizon predictions can be sensitive to model
assumptions about trend, seasonality, and autoregressive dependence.

# Further Reading

The examples above illustrate the basic univariate workflow for fitting,
diagnosing, visualizing, and making short-term predictions from a `tempssm`
model. The detailed manual extends this workflow to exogenous variables,
additional model diagnostics, and time-series cross-validation:

https://github.com/akihirao/tempssm/blob/main/tools/manual/tempssm_manual.pdf
