---
title: "tempssm Reference Manual"
subtitle: "Version `r packageVersion('tempssm')`"
author: "Akira S. Hirao, Shinya Baba, and Momoko Ichinokawa"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm+tex_math_dollars
vignette: >
  %\VignetteIndexEntry{tempssm Reference Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Summary

`tempssm` is an R package for state-space modeling of environmental
temperature time series, including air and water temperature observations. It
provides a practical framework for assessing how long-term trend, seasonal
variation, autoregressive dependence, and optional exogenous effects contribute
to observed temporal variation. The package facilitates the application of
linear Gaussian state-space models estimated by Kalman filtering and smoothing,
using the `KFAS` package as the computational backend (Helske, 2017).

# Key Features

- Fits linear Gaussian state-space models to environmental temperature time
  series.
- Represents temperature dynamics using interpretable latent components:
  long-term trend, seasonal variation, autoregressive dependence, and optional
  exogenous effects.
- Supports arbitrary seasonal frequencies, while the current examples
  and validation focus primarily on monthly temperature data.
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1)).
- Includes time-series cross-validation tools for model evaluation.

# Input Data Format

The core modeling functions in `tempssm` expect input time series to be
supplied as R `ts` objects. The `ts` class is base R's standard format for
regularly spaced time series (see the `stats::ts` documentation:
<https://search.r-project.org/R/refmans/stats/html/ts.html>). The package also
provides utility functions for converting common tabular and observational data
formats into `ts` objects before model fitting.

The seasonal cycle used by `tempssm()` is taken from the `frequency` attribute
of the input `ts` object. For example, `frequency = 12` represents monthly
data.

# Prior Art and Scope

`tempssm` provides a domain-focused workflow for analyzing temperature time
series with linear Gaussian state-space models. It brings together model
construction, component extraction, uncertainty summaries, residual
diagnostics, visualization, and time-series cross-validation in a single R
package interface tailored to temperature applications.

The package builds on established statistical methodology, including linear
Gaussian state-space modeling, Kalman filtering, and Kalman smoothing. Model
estimation is handled through the `KFAS` package, which provides a general
framework for state-space models in R.

The initial implementation was adapted from the supplementary code provided by
Baba (2024), accompanying Baba et al. (2024), which analyzed sea temperature
trends using a linear Gaussian state-space model. The supplementary code is
publicly available at:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

Compared with that prior implementation, `tempssm` extends the workflow into a
reusable R package interface with input validation, documented S3 methods,
tests, diagnostics, cross-validation utilities, and examples for broader
temperature time-series analysis.

The next section gives a brief model overview. The mathematical formulation is
described separately in the package vignette `model-specification`.

# Model Overview

The model implemented in `tempssm` can be viewed as an extension of the Basic
Structural Time Series Model (BSTSM), a standard state-space formulation that
represents an observed time series using latent trend, seasonal, and irregular
components. Following the temperature time-series application of Baba et al.
(2024), `tempssm` keeps this interpretable decomposition and adds
autoregressive dependence and optional exogenous effects. This makes it useful
for separating long-term temperature change, seasonal cycles, and short-term
departures.

The computational backend is `KFAS`, and the fitted `tempssm` object stores
both filtering and smoothing estimates. Unless otherwise stated, this manual
interprets summary statistics, diagnostics, and plots based on smoothed state
estimates.

The detailed mathematical formulation is separated from this tutorial-style
manual to reduce duplication and improve maintainability. See the package
model specification document for the observation equation, state equations,
seasonal constraint, autoregressive and exogenous components, parameter count,
and estimation procedure. It is available as a PDF in the package repository:

<https://github.com/akihirao/tempssm/blob/main/vignettes/model-specification.pdf>

# How to Use

## Set Environment

Load the following R packages to run the examples below. If any packages are
not installed, please install them as needed.

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
library(forecast)

library(purrr) 
library(tibble)
library(readr)
library(dplyr)
library(ggplot2)

library(patchwork)
```

## Exercise I: Applying a State-Space Model to a Univariate Temperature Time Series

### Objective

In Exercise I, we apply a linear Gaussian state-space model to a univariate
temperature time series. This exercise introduces basic modeling without
exogenous variables and examines the role of autoregressive dynamics.

### Loading the Sea Surface Temperature (SST) Dataset

A sample sea surface temperature (SST) dataset is included in the package.

- **Dataset**: Monthly sea surface temperature (SST) off Yamaguchi Pref., Japan  
- **Unit**: °C  
- **Period**: February 2002 to December 2023

This dataset was created by aggregating daily SST data obtained from the Japan
Meteorological Agency website (https://www.jma.go.jp/jma/indexe.html) into
monthly values.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(yamaguchi_sst) # load a ts object of SST off Niigata
head(yamaguchi_sst)
summary(yamaguchi_sst)
```

### Plotting the Monthly SST Time Series

We begin by visualizing the monthly SST time series to examine its
overall structure, including apparent trends, seasonal variability,
and the presence or absence of missing observations.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_yamaguchi_sst <- forecast::autoplot(yamaguchi_sst) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)") +
  ggtitle("Monthly SST off Yamaguchi, Japan") +
  theme_classic()

plot(plt_yamaguchi_sst)
```

```{r echo = FALSE}
yamaguchi_sst_mean <- mean(yamaguchi_sst, na.rm = TRUE)
```

The overall mean SST is approximately
`r sprintf("%.1f", yamaguchi_sst_mean)` °C, and a clear seasonal pattern is
visible. The series contains no missing observations. Although SST appears to
increase over the study period, year-to-year variability is also evident,
making the long-term trend difficult to assess from the raw time series alone.

### Applying a Linear Gaussian State-Space Model

When a `ts` object containing temperature time-series data (here,
`yamaguchi_sst`) is passed to the core function `tempssm()`, model construction
and parameter estimation are performed together. The returned S3 object of
class `tempssm` (here, `res_ar1`) stores the filtering and smoothing estimates,
as well as the constructed model and input data. By default, `tempssm()` fits a
first-order autoregressive model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# model with first-order autoregressive component
res_ar1 <- tempssm(yamaguchi_sst) # (ar_order=1: default)
summary(res_ar1)
```

From the summary output, confirm that the model has converged
(`Converged: TRUE`). The output also reports statistics such as the number of
parameters (`k`), the log-likelihood, the likelihood type, and the number of
diffuse initial states. The parameter estimates include the observation error
variance (`H`), the process error variance of the long-term trend component
(`Q trend`), the process error variance of the seasonal component
(`Q season`), the process error variance of the autoregressive component, and
the first-order autoregressive coefficient (`AR1`).

The log-likelihood and the associated number of estimated parameters can also
be extracted directly from the fitted `tempssm` object using `logLik()`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
ll <- logLik(res_ar1)
ll
attr(ll, "df") # number of parameters
```

By default, `tempssm()` uses the KFAS marginal likelihood for parameter
estimation. The diffuse likelihood remains available by explicitly setting
`marginal = FALSE`. The selected likelihood type is stored in the fitted
`tempssm` object and is used consistently by downstream methods such as
`logLik()` and `summary()`.

The package intentionally does not compute AIC for `tempssm` objects. The
log-likelihood and parameter count remain available through `logLik()` for
users who need them for their own model-assessment workflows.

### Plotting Long-Term Trend, Drift, Seasonal, and Autoregressive Components

We extract the corresponding latent components from the state-space model and
visualize the estimated long-term evolution of temperature levels and their
rate of change (drift). Plotting seasonal variation and autoregressive
dependence as well makes the underlying trend structure easier to examine.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot all components at once
plot(res_ar1)
```

```{r echo = FALSE}
mean_drift_year <- mean(get_drift_ts(res_ar1))
```

The long-term trend in the upper-left panel indicates an increasing SST pattern
over the study period. In this example, the average annual rate of SST increase
over the study period is approximately
`r sprintf("%.3f", mean_drift_year)` °C.

However, the rate of SST change itself also appears to vary over time. Focusing
on the rate of change (drift) in the upper-right panel, the rate of change
declines to zero or near zero during the 2000s and then increases after 2010.
The shaded gray areas represent 95% confidence intervals for the estimated
latent states and illustrate the uncertainty associated with each estimated
component.

The standard plotting interface is `plot(res)`. The ggplot2-style interface
`autoplot(res)` is also available and produces the same component plot by
default. To obtain a ggplot object for an individual component, specify the
`component` argument in `autoplot()` as follows.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plt_level <- autoplot(res_ar1, component = c("level"))
plt_drift <- autoplot(res_ar1, component = c("drift"))
plt_season <- autoplot(res_ar1, component = c("season"))
plt_ar <- autoplot(res_ar1, component = c("ar"))
```

### Model Diagnostics

The package provides diagnostic tools for checking whether the fitted model has
left notable structure in the residuals. In particular, residual time-series
plots, residual autocorrelation, residual distributions, and Ljung-Box tests
can be used to assess remaining temporal dependence and departures from the
Gaussian error assumption.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plot_tempssm_residual_diagnostics(res_ar1)
```

In the model diagnostic plot, the upper panel shows the residual time series,
the lower-left panel shows the residual autocorrelation plot (ACF plot), and
the lower-right panel shows the residual frequency distribution. These plots
should be checked for any notable residual patterns. 

In this example, the ACF plot shows an isolated spike around lag 5 and lag 27, 
but there is no clear sequence of successive significant lags or repeated 
seasonal pattern. Such isolated spikes can occur when many lags are inspected, 
so they should be interpreted together with the Ljung-Box test and the residual 
time-series plot rather than used as the sole basis for changing the model.　

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag <- diagnose_residuals(res_ar1)
print(diag)
```

The `lb_stat`, `lb_lag`, and `lb_pvalue` columns returned by
`diagnose_residuals()` correspond to the Ljung-Box test statistic, the lag used
in the test, and the corresponding P-value. For monthly time series, the
default Ljung-Box lag is the seasonal frequency, so the default test evaluates
residual autocorrelation up to lag 12. In this example, the Ljung-Box test does
not indicate significant residual autocorrelation up to lag 12.

If residual autocorrelation at longer lags is of concern, the lag can be
specified manually. For example, the following code repeats the Ljung-Box test
up to lags 24 and 36.

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag_lag24 <- diagnose_residuals(res_ar1, lb_lag = 24)
diag_lag36 <- diagnose_residuals(res_ar1, lb_lag = 36)

diag_lags <- rbind(diag, diag_lag24, diag_lag36)
diag_lags$check <- c("lag 12", "lag 24", "lag 36")
diag_lags[, c("check", "lb_stat", "lb_lag", "lb_pvalue", "kurtosis")]
```

These diagnostics should be interpreted together with the residual ACF plot.
A small number of isolated ACF spikes can occur by chance when many lags are
inspected, whereas repeated or periodic spikes may suggest remaining temporal
structure.

### Examining the Autoregressive (AR) Order

In this manual, the default AR(1) specification is used as the working model.
The autoregressive component is included to absorb short-term serial
dependence that is not represented by the latent level, drift, and seasonal
components. A higher AR order should therefore not be treated as an automatic
improvement. Instead, the AR order can be adjusted when residual diagnostics
suggest that notable autocorrelation remains.

For example, if residual autocorrelation remains after fitting the AR(1)
model, users can fit an AR(2) model and repeat the same diagnostics.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Optional sensitivity check with a second-order autoregressive component
res_ar2 <- tempssm(yamaguchi_sst, ar_order = 2)
summary(res_ar2)
plot_tempssm_residual_diagnostics(res_ar2)
diag_lag12_ar2 <- diagnose_residuals(res_ar2, lb_lag = 12)
diag_lag24_ar2 <- diagnose_residuals(res_ar2, lb_lag = 24)
diag_lag36_ar2 <- diagnose_residuals(res_ar2, lb_lag = 36)
diag_lags_ar2 <- rbind(diag_lag12_ar2, diag_lag24_ar2, diag_lag36_ar2)
diag_lags_ar2$check <- c("lag 12", "lag 24", "lag 36")
diag_lags_ar2[, c("check", "lb_stat", "lb_lag", "lb_pvalue", "kurtosis")]
```

In a sensitivity check with an AR(2) specification, the short-lag residual
autocorrelation was also limited, but the residual ACF showed a negative spike
around lag 24, 26, 27. This pattern suggests that increasing the AR order 
does not necessarily improve all aspects of the residual structure. Therefore, 
the AR(2) model should be treated as a model requiring additional diagnostic
attention rather than as an automatic improvement over the AR(1) baseline.

In the present example, the residual diagnostics do not provide strong
evidence that a higher AR order is needed, so we proceed with the AR(1) model
as the baseline specification.

### Estimated Parameters and Latent-State Components

The long-term trend component and its rate of change (drift) can be extracted
as `ts` objects as follows.
r
```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res_ar1$kfs$alphahat
head(alpha_hat)

#　Smoothing estimate of level component
level_ts <- get_level_ts(res_ar1)

#　Smoothing estimate of drift component
drift_ts <- get_drift_ts(res_ar1)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts) 
print(mean_drift_year)
```
 
The average annual rate of SST increase was estimated to be
`r sprintf("%.4f", mean_drift_year)` °C.

### Short-Term Prediction

A fitted `tempssm` object can also be passed to `predict()` to obtain
short-term predictions. By default, `predict(res)` returns a one-step-ahead
prediction beyond the end of the observed series. This is useful for visual
checks of how the fitted model extrapolates the estimated level, seasonal, and
autoregressive components.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_1 <- predict(res_ar1)
pred_1
```

Predictions for multiple future time points can be requested by setting the
`n.ahead` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12 <- predict(res_ar1, n.ahead = 12)
pred_12
```

Prediction uncertainty can also be returned by setting the `interval`
argument. With `interval = "confidence"`, the returned lower and upper bounds
describe uncertainty in the predicted mean response. With
`interval = "prediction"`, the bounds describe uncertainty in a future
observation and therefore also include observation error; prediction intervals
are usually wider than confidence intervals. The confidence level is controlled
by the `level` argument and defaults to 0.95.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12_pi <- predict(
  res_ar1,
  n.ahead = 12,
  interval = "prediction",
  level = 0.95
)

pred_12_pi
```

These predictions should be interpreted as model-based extrapolations rather
than definitive forecasts. Uncertainty generally increases as the prediction
horizon becomes longer, and long-horizon predictions can be sensitive to model
assumptions about trend, seasonality, and autoregressive dependence.


## Exercise II: Applying a State-Space Model to a Temperature Time Series with an Exogenous Variable

### Objective

We extend the state-space modeling framework to examine the effect of an
exogenous factor on temperature variation. Specifically, we examine the effect
of the Pacific Decadal Oscillation (PDO) as an exogenous variable on SST
observed off Yamaguchi Prefecture, Japan.

### Loading the PDO Index Dataset: PDO Index as an Exogenous Variable

- **Data**: Monthly Pacific Decadal Oscillation (PDO) index (JMA)  
- **Period**: January 1901 to December 2025

The Pacific Decadal Oscillation (PDO) index is defined as the projection of 
monthly mean sea surface temperature (SST) anomalies onto the leading empirical 
orthogonal function (EOF) of SST variability over the North Pacific north of 
20°N. The EOF is computed using SST anomalies for 1901–2000, defined relative 
to the 1901–2000 monthly climatology. To remove the global warming signal, 
the global-mean SST anomaly is subtracted from each grid point prior to 
the EOF analysis. In this package, we use the PDO index provided by 
the Japan Meteorological Agency (JMA), available at
<https://www.data.jma.go.jp/kaiyou/data/shindan/b_1/pdo/pdo.txt>.


```{r message = FALSE, warning = FALSE, echo = TRUE}
data(pdo) # load a ts object of PDO index
head(pdo)
```

### Intersecting the Temperature and PDO Time Series

For state-space modeling with exogenous variables, all input time series must
share a common and aligned time index. In this step, the temperature and PDO
time series are restricted to their overlapping period by trimming the leading
and trailing portions so that both datasets cover the same time span.

The function `tempssm::trim_ts_overlap()` is used to align the two `ts` objects
on a shared timeline and returns time series containing only the common period.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate an object on a shared timeline
trimmed_series <- trim_ts_overlap(yamaguchi_sst,
                                  pdo,
                                  temp_name = "Temp",
                                  exo_name = "PDO")

yamaguchi_sst_trim <- trimmed_series$temperature
pdo_trim <- trimmed_series$exogenous

start(yamaguchi_sst_trim)
end(yamaguchi_sst_trim)

start(pdo_trim)
end(pdo_trim)

```

### Plotting Time Series of SST and the PDO Index

We visualize the time-series structures of SST and the PDO index.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_yamaguchi_sst_trim <- forecast::autoplot(yamaguchi_sst_trim) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)") +
  ggtitle("Monthly SST off Yamaguchi, Japan") +
  theme_classic()


plt_pdo <- forecast::autoplot(pdo_trim) +
  labs(x = "Time (year)", y = "PDO index") +
  ggtitle("PDO index") +
  theme_classic()

plt_yamaguchi_sst_trim + plt_pdo + patchwork::plot_layout(ncol=1)

```

The PDO index data used in this exercise contain no missing values. When using
your own exogenous-variable data, however, make sure that the exogenous
variables do not contain missing values before fitting the model. `tempssm()`
does not allow missing values in exogenous variables, and model fitting will
stop with an error if such data are supplied. If necessary, apply appropriate
preprocessing, such as imputation, before the analysis.

In contrast, missing values in the dependent temperature time series are
allowed. In the state-space model, these missing values are treated as
unobserved responses, and model estimation proceeds through Kalman filtering
and smoothing.

### Applying a Model Without an Exogenous Variable

We first apply a baseline state-space model that does not include any
exogenous variables. This model serves as a reference case in which temperature
variation is explained solely by the latent long-term trend, seasonal cycle,
and autoregressive dependence. Although the same model was already fitted in
Exercise I, here we explicitly specify the dependent-variable argument and
store the returned object as `res_without`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_without <- tempssm(temp_data = yamaguchi_sst_trim, ar_order = 1) 
summary(res_without)

plot_tempssm_residual_diagnostics(res_without)
diag_res_without <- diagnose_residuals(res_without)
print(diag_res_without)
```

The residual diagnostics for this AR(1) baseline model do not indicate
significant residual autocorrelation up to lag 12. This result suggests that
the default first-order autoregressive structure is a reasonable starting
point for the model without the exogenous variable.

This baseline model provides a useful benchmark for evaluating the additional
explanatory power of the PDO index introduced in the next section.

### Applying a Model With an Exogenous Variable

We construct a model that uses the PDO index as a univariate exogenous
variable. In the `tempssm()` function, the PDO data (`pdo_trim`) are supplied
to the exogenous-variable argument (`exo_data`). The temperature time series
(`yamaguchi_sst_trim`) is supplied to the dependent-variable argument
(`temp_data`), which was implicit in the previous examples.

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_with <- tempssm(temp_data = yamaguchi_sst_trim,
                    exo_data = pdo_trim,
                    ar_order = 1) 
summary(res_with)

plot_tempssm_residual_diagnostics(res_with)
diag_res_with <- diagnose_residuals(res_with)
print(diag_res_with)
```

```{r echo = FALSE}
exo_coef <- get_exo_coef(res_with)
pdo_coef <- exo_coef$Coefficient[exo_coef$Variable == "PDO"]
pdo_lwr <- exo_coef$lwr[exo_coef$Variable == "PDO"]
pdo_upr <- exo_coef$upr[exo_coef$Variable == "PDO"]
```

The residual diagnostics for the model with the PDO index also do not indicate
significant residual autocorrelation up to lag 12. Thus, for both models
considered here, the AR(1) specification provides an adequate working
autoregressive structure according to these residual checks.

We then examine the model estimates. The estimated coefficient for the
exogenous PDO index is negative (`r sprintf("%.2f", pdo_coef)`), and its 95%
confidence interval does not include zero, ranging from
`r sprintf("%.2f", pdo_lwr)` to `r sprintf("%.2f", pdo_upr)`. This indicates a
statistically significant negative relationship between local SST off Yamaguchi
Prefecture and PDO variability.

Specifically, after accounting for the underlying long-term trend, seasonal
cycle, and autoregressive dependence, the model suggests that a one-unit
increase in the PDO index is associated with an average decrease of
approximately `r sprintf("%.2f", abs(pdo_coef))` °C in monthly SST. This result
implies that positive phases of the PDO systematically contribute to cooler
conditions at the study site.

Importantly, the effect of this exogenous variable is identified as an
additional factor beyond the internal dynamics of the temperature time series,
rather than as an artifact of improved representation of the long-term trend or
temporal dependence. Compared with the baseline model without exogenous
variables, the statistical significance of the PDO coefficient indicates that
the PDO acts as an independent large-scale climate driver influencing local
temperature variation.

For models with exogenous variables, future values of the exogenous variables
are needed when calling `predict()`. If such values are available, they can be
passed through `new_exo_data`. For a simple one-step-ahead visual check,
`exo_strategy = "last"` carries the final observed exogenous value forward.
This is a persistence assumption for the exogenous variable, not a separate
forecast model for the PDO index.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_with_last_exo <- predict(res_with, exo_strategy = "last")
pred_with_last_exo
```

### Time-Series Cross-Validation (tsCV)

Time-series cross-validation (tsCV) evaluates model performance based on
out-of-sample prediction errors while respecting the temporal ordering of the
data. It therefore provides an additional and independent perspective on model
adequacy.

In time-series cross-validation, the model is repeatedly fitted to expanding or
sliding training windows, and predictive performance is evaluated on subsequent
observations. Unlike random cross-validation, this procedure prevents
information leakage from the future to the past and is therefore suitable for
model validation with temporal data.

Here, by comparing cross-validation metrics for models with and without the
exogenous PDO variable, we assess whether the statistical signal found in the
PDO coefficient is also reflected in out-of-sample predictive performance.

For the model with the PDO index, the test-period PDO values are supplied to
the forecast step as exogenous variables. Therefore, this tsCV setting evaluates
conditional forecasts, or hindcasts, under the assumption that the PDO values
during each test period are known. This is appropriate when the aim is to assess
the explanatory or reconstructive value of the PDO index.

For real-time future forecasting, however, future PDO values are usually
unknown. In that setting, users must either provide future PDO scenarios,
forecast the PDO index with a separate model, or use a simplified assumption
such as carrying the last observed value forward. The results below should
therefore be interpreted as evidence for improved conditional predictive
performance, not as a direct guarantee of operational forecast improvement.


```{r message = FALSE, warning = FALSE, echo = TRUE}

# (Optional) Load packages for parallel processing.
# These are only required if you want to enable parallel execution:
# - future: controls how parallel workers are launched
# - future.apply: provides parallelized apply functions
# - progressr: optional progress bar support
library(future.apply)
library(future)
library(progressr)

if (interactive()) {
  future::plan(multisession)
} else {
  future::plan(sequential)
}


# ts cross-validation of the model without exogenous variables

## Generate a list of training and test datasets with their indices
# Procedure for constructing year-based time-series cross-validation folds:
# 
# First training data: January 1982–December 2017;
# First test data: one year starting from January 2018
# 
# Second training data: January 1983–December 2018;
# Second test data: one year starting from January 2019
# ...
# 
# Eighth training data: January 1989–December 2024;
# Eighth test data: one year starting from January 2025
#
# These folds are automatically generated by the ts_train_test_split() function.

# Generate training and test dataset for model without PDO index
folds_without <- ts_train_test_split(
  temp_data = yamaguchi_sst_trim,
  exo_data = NULL,
  initial = 432, # 432 monthly observations from Jan 1982 to Dec 2017
  horizon = 12, # forecast 12 monthly time-series
  step = 12, # training data is moved in one-year steps
  fixed_window = TRUE,
  allow_partial = FALSE
  )

# Generate training and test dataset for model with PDO index
folds_with <- ts_train_test_split(
  temp_data = yamaguchi_sst_trim,
  exo_data = pdo_trim,
  initial = 432, # 432 monthly observations from Jan 1982 to Dec 2017
  horizon = 12, # forecast 12 monthly time-series
  step = 12, # training data is moved in one-year steps
  fixed_window = TRUE,
  allow_partial = FALSE
  )


# *****************************************
# Executing time-series cross validation

#-------------------------------------------------- 
# Model without the exogenous variable of PDO index

# Single processing
# cv_without_results <- ts_cv_run(folds_without, ar_order = 1, use_season = TRUE)

# Parallel processing
cv_without_results <- future.apply::future_lapply(
  folds_without,
  ts_cv_run_fold,
  ar_order = 1,
  use_season = TRUE,
  future.seed = TRUE
)

# Computing assessment indexes
metrics_without <- lapply(cv_without_results, compute_cv_metrics)

# tidy summary
cv_without_tbl <- ts_cv_collect(cv_without_results, metrics_without) %>%
  mutate(Model="Without")


#-------------------------------------------------- 
# Model with the exogenous variable of PDO index

# Single processing
# cv_with_results <- ts_cv_run(folds_with, ar_order = 1, use_season = TRUE)

# Parallel processing
cv_with_results <- future.apply::future_lapply(
  folds_with,
  ts_cv_run_fold,
  ar_order = 1,
  use_season = TRUE,
  future.seed = TRUE
)

# Computing assessment indexes
metrics_with <- lapply(cv_with_results, compute_cv_metrics)

# tidy summary
cv_with_tbl <- ts_cv_collect(cv_with_results, metrics_with) %>%
  mutate(Model="With")


cv_tbl <- bind_rows(cv_without_tbl, cv_with_tbl)

cv_comparison <- compare_ts_cv(
  list(
    Without = cv_without_tbl,
    With = cv_with_tbl
  )
)

cv_comparison %>% knitr::kable()


plt_MAE <- ggplot(data=cv_tbl,
                  aes(x=Model,y=MAE)) +
  geom_boxplot()

plt_MASE_naive <- ggplot(data=cv_tbl,
                   aes(x=Model,y=MASE_naive)) +
  geom_boxplot()

plt_MASE_seasonal <- ggplot(data=cv_tbl,
                   aes(x=Model,y=MASE_seasonal)) +
  geom_boxplot()


plt_tsCV <- plt_MAE + plt_MASE_naive + plt_MASE_seasonal + patchwork::plot_layout(nrow=1)

plot(plt_tsCV)

```

```{r echo = FALSE}
monthly_mean_sst <- tapply(
  as.numeric(yamaguchi_sst_trim),
  cycle(yamaguchi_sst_trim),
  mean,
  na.rm = TRUE
)
seasonal_amplitude <- diff(range(monthly_mean_sst, na.rm = TRUE))
mae_without <- cv_comparison$mean_MAE[cv_comparison$model == "Without"]
mae_with <- cv_comparison$mean_MAE[cv_comparison$model == "With"]
mae_relative_to_season <- 100 * mae_with / seasonal_amplitude
mae_improvement <- 100 * (mae_without - mae_with) / mae_without
n_folds <- unique(cv_comparison$n_folds)
n_folds <- if (length(n_folds) == 1L) n_folds else min(n_folds)
```

The tsCV results show that the model with the exogenous variable performs
better than the model without it. The comparison table produced by
`compare_ts_cv()` summarizes the number of folds, convergence rates, and mean
prediction-error metrics for each model, while the boxplots show how the
fold-level errors vary across validation periods. In this tutorial, the number
of tsCV iterations is set to `r n_folds` to reduce execution time. In actual
validation, please use a sufficient number of iterations.

The mean MAE of the model with the exogenous variable is approximately
`r sprintf("%.2f", mae_with)` °C. Relative to the seasonal amplitude based on
monthly mean SST (about `r sprintf("%.1f", seasonal_amplitude)` °C), this
corresponds to roughly `r sprintf("%.1f", mae_relative_to_season)`%. Thus, the
prediction error is small compared with the dominant seasonal scale of
variation in this data set, and may be practically useful for short-term marine
monitoring.

Although the number of validation folds is limited to `r n_folds`, the model
with the exogenous variable reduces the mean MAE by approximately
`r sprintf("%.1f", mae_improvement)`% compared with the model without the
exogenous variable. Similar improvements are also seen in the MASE metrics.
This interpretation should be kept conditional on the tsCV setting used here,
in which observed PDO values during the test periods are supplied as exogenous
variables.

Taken together, the results consistently support including the PDO index as an
exogenous variable in the state-space model. The PDO coefficient is
statistically significant, indicating a clear effect on temperature variation.
Time-series cross-validation further indicates that this effect is accompanied
by better out-of-sample predictive performance.

These complementary lines of evidence provide robust and multifaceted support
for adopting the exogenous-variable model.

Although this exercise used a single exogenous variable, `tempssm()` also
supports models with multiple exogenous variables supplied as multivariate
time-series data. Users interested in broader applications can extend the same
workflow to other environmental, climatic, or ecological covariates. In doing
so, model structure should be considered carefully for the target data, while
checking model diagnostics and out-of-sample predictive performance.


### Comparing Estimation Patterns Between Models With and Without the Exogenous Variable

We compare how the estimated patterns of the long-term trend and its rate of
change differ between the models with and without the exogenous variable.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_without_ts <- autoplot(res_without,
                                 component=c("level")
                                 ) +
  labs(title="Model without the PDO index") +
  theme_classic()

plt_drift_without_ts <- autoplot(res_without, 
                             component=c("drift")
                             ) + 
  labs(title="Model without the PDO index") +
  theme_classic()

plt_level_drift_without_ts <- plt_level_without_ts + 
  plt_drift_without_ts + 
  patchwork::plot_layout(ncol=1)

plot(plt_level_drift_without_ts)


plt_level_with_ts <- autoplot(res_with,
                          component=c("level")
                          )+ 
  labs(title="Model with the PDO index") +
  theme_classic()

plt_drift_with_ts <- autoplot(res_with,
                          component=c("drift")
                          ) + 
  labs(title="Model with the PDO index") + 
  theme_classic()

plt_level_drift_with_ts <- plt_level_with_ts + 
  plt_drift_with_ts + 
  patchwork::plot_layout(ncol=1)

plot(plt_level_drift_with_ts)

#　Smoothing estimate of drift component
mean_drift_year_without <- get_drift_ts(res_without) %>%
  mean()
print(mean_drift_year_without)

mean_drift_year_with <- get_drift_ts(res_with) %>%
  mean()
print(mean_drift_year_with)

```
The gray areas in the plots above show 95% confidence intervals.

In the model without an exogenous variable, the plots of the long-term trend
and its rate of change show a wavelike pattern. In contrast, in the model with
the exogenous variable, the long-term trend and its rate of change show a more
stable pattern. This suggests that the wavelike pattern reflects the influence
of the exogenous PDO variable.

The estimated annual rate of SST change over the observation period is
`r sprintf("%.4f", mean_drift_year_without)` in the model without the
exogenous variable and `r sprintf("%.4f", mean_drift_year_with)` in the model
with the exogenous variable.

In this case, the model incorporating the PDO index is considered preferable
for accurately estimating the SST trend. Thus, the SST warming rate, which was
estimated somewhat higher in the model without the exogenous variable, is
estimated more conservatively after introducing the model with the exogenous
variable.


# Using Your Own Data

## Quick Example

`my_ts` is a user-supplied `ts` object containing temperature time-series data.
It can be passed directly to `tempssm()`.

```{r, eval = FALSE}
## example of ts object (not run)
## my_ts <- ts(rnorm(100), frequency = 12)
res <- tempssm(my_ts)
```

## Preparing Your Own Dataset

### CSV Format
Prepare a CSV file of monthly temperature time series with the following
columns:
- Year
- Month
- Temp

Example:
```text
Year,Month,Temp
2010,8,13.6
2010,9,6.8
2010,10,NA
2010,11,-1.4
...
```

* Use NA for missing temperature values, and always keep the corresponding Year
  and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.


### Converting CSV to a Time Series

An example CSV file included in this package is available in inst/extdata. The
example dataset contains monthly air temperature observations from Mt. Akadake,
Hokkaido, Japan (1,840 m). The original data are available from the Monitoring
Sites 1000 Project of the Ministry of the Environment of Japan (KOZ01.zip,
https://www.biodic.go.jp/moni1000/findings/data/index.html).

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")
akadake_temp <- tempssm::read_monthly_temp_ts(path)
head(akadake_temp)
```

The `read_monthly_temp_ts()` function reads this type of CSV file and converts
it into an R `ts` object for use with `tempssm()`. If finer control is needed,
you can also construct a `ts` object manually; see the `stats::ts`
documentation for details:
<https://search.r-project.org/R/refmans/stats/html/ts.html>.



# References

The statistical modeling framework implemented in `tempssm` is based on the
methodology described in Baba et al. (2024), and the accompanying supplementary
materials and code repository served as the initial source for development of
this package.

Baba, S., Ishii, H., and Yoshiyama, T. (2024).
Estimating sea temperature trends using a linear Gaussian state-space model
in Jogashima, Kanagawa, Japan.
*Bulletin of the Japanese Society of Fisheries Oceanography*,
88(3), 190–199.
(In Japanese with an English abstract.)
https://doi.org/10.34423/jsfo.88.3_190

Baba, S. (2024).
Supplementary code and test data for estimating sea temperature trends using a
linear Gaussian state-space model.
GitHub repository:
https://github.com/logics-of-blue/sea-temperature-trend-jogashima

Helske, J. (2017).
KFAS: Exponential Family State Space Models in R.
*Journal of Statistical Software*, 78(10), 1–39.
https://doi.org/10.18637/jss.v078.i10

# Appendix: Utility Functions

The following utility functions are provided to support data preparation and
exploratory analysis.


## 1. `read_monthly_temp_ts()`

The `read_monthly_temp_ts()` function converts monthly temperature data stored
in a CSV file into an R `ts` object. By enforcing a simple and consistent data
format, it is intended to make externally prepared time-series data easier to
import into tempssm.

### Example

For monthly temperature data, prepare a CSV file with a header row. By default,
the column names must be `Year`, `Month`, and `Temp`, where `Temp` represents
the observed temperature value.

```text
Year,Month,Temp
2001,1,10.4
2001,2,8.2
2001,3,NA
2001,4,13.6
2001,5,16.1
...
```

* Use NA for missing temperature values, and always keep the corresponding Year
  and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.

The following example uses the sample CSV file included in the package. It
contains monthly air temperature observations from Mt. Akadake, Hokkaido,
Japan.

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")

# Read the CSV file and convert it to a monthly ts object
temp_ts <- read_monthly_temp_ts(path)
head(temp_ts)
```

## 2. `convert_monthly_df_to_ts()`

The `convert_monthly_df_to_ts()` function converts a data frame containing
monthly temperature data into an R `ts` object. It is designed to support
workflows in which temperature data are first imported or prepared as a data
frame and then used for time-series analysis.

The input data frame is expected to contain at least two columns: a date column
(`Date`) and a temperature column (`Temp`). For the date column (`Date`), assign
an arbitrary day of the month, such as the first day, so that the dates
correspond to monthly aggregated temperature values.

### Example

```{r message = FALSE, warning = FALSE}
# Create a data frame of monthly temperature data
df <- data.frame(
  Date = as.Date(c(
    "2001-01-01",
    "2001-02-01",
    "2001-03-01",
    "2001-04-01",
    "2001-05-01")
    ),
  Temp = c(10.4, 8.2, NA, 13.6, 16.1)
  )

# Convert to a monthly ts object
temp_ts <- convert_monthly_df_to_ts(df)
head(temp_ts)
```

## 3. `get_jma_sst_ts()`

The `get_jma_sst_ts()` function downloads daily mean sea surface temperature
(SST) data for Japanese coastal waters provided by the Japan Meteorological
Agency (JMA). It aggregates the daily values into monthly means and returns the
resulting time series as an object of class `ts`.

### Example

The following example demonstrates how to download SST data for the southern
coastal waters of Ibaraki Prefecture, Japan.

The argument `sea_area_id` specifies the numeric identifier of a sea area
defined by JMA. For example, `138` corresponds to the coastal waters off
southern Ibaraki. A list of available sea area IDs and their corresponding
regions is provided by JMA at:

<https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html>

```{r message = FALSE, warning = FALSE, echo = TRUE}
sst_138_ts <- get_jma_sst_ts(sea_area_id = 138)
head(sst_138_ts)
```

## 4. `compute_temp_anomaly()`

The `compute_temp_anomaly()` function computes anomalies from a time series
provided as an R `ts` object. Anomalies are calculated by subtracting the
corresponding long-term mean for each periodic unit, such as month, from each
observation. This transformation is useful for examining interannual
variability after removing periodic effects.

The reference period used to compute anomalies can be specified via a function
argument. By default, the climatology is computed over the entire available
time series (see `?compute_temp_anomaly`).

This transformation is useful for exploratory analysis and modeling
applications that focus on departures from typical seasonal conditions.

### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
data(niigata_sst)
niigata_sst_anomaly <- compute_temp_anomaly(niigata_sst)
plt_niigata_sst_anomaly <- forecast::autoplot(niigata_sst_anomaly) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)")
plot(plt_niigata_sst_anomaly) 
```

## 5. `compute_monthly_climatology()`

The `compute_monthly_climatology()` function computes the climatological mean
seasonal cycle from a `ts` object by averaging each seasonal value across
years. It is primarily intended for exploratory analysis and visualization of
seasonal structure in temperature time series.

### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(niigata_sst)
monthly_seasonal_cycle_niigata_sst <- compute_monthly_climatology(niigata_sst) 
summary(monthly_seasonal_cycle_niigata_sst)

plt_monthly_seasonal_cycle_niigata_sst <- ggplot(
  data = monthly_seasonal_cycle_niigata_sst,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temp.~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  ) +
  theme_classic()


plot(plt_monthly_seasonal_cycle_niigata_sst)
```
