---
title: "tempssm Reference Manual"
subtitle: "Version `r packageVersion('tempssm')`"
author: "Akira S. Hirao, Shinya Baba, and Momoko Ichinokawa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
#    toc: true
#  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm+tex_math_dollars
vignette: >
  %\VignetteIndexEntry{tempssm Reference Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Summary

`tempssm` is an R package for state-space modeling of environmental
temperature time series, including air and water temperature observations. It
provides a practical framework for assessing how long-term trend, seasonal
variation, autoregressive dependence, and optional exogenous effects contribute
to observed temporal variation. The package facilitates the application of
linear Gaussian state-space models estimated by Kalman filtering and smoothing,
using the `KFAS` package as the computational backend (Helske, 2017).

# Key Features

- Fits linear Gaussian state-space models to environmental temperature time
  series.
- Represents temperature dynamics using interpretable latent components:
  long-term trend, seasonal variation, autoregressive dependence, and optional
  exogenous effects.
- Supports arbitrary seasonal frequencies, while the current examples
  and validation focus primarily on monthly temperature data.
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1)).
- Includes time-series cross-validation tools for model evaluation.

# Input Data Format

The core modeling functions in `tempssm` expect input time series to be
supplied as R `ts` objects. The `ts` class is base R's standard format for
regularly spaced time series (see the `stats::ts` documentation:
<https://search.r-project.org/R/refmans/stats/html/ts.html>). The package also
provides utility functions for converting common tabular and observational data
formats into `ts` objects before model fitting.

The seasonal cycle used by `tempssm()` is taken from the `frequency` attribute
of the input `ts` object. For example, `frequency = 12` represents monthly
data.

# Prior Art and Scope

`tempssm` provides a domain-focused workflow for analyzing temperature time
series with linear Gaussian state-space models. It brings together model
construction, component extraction, uncertainty summaries, residual
diagnostics, visualization, and time-series cross-validation in a single R
package interface tailored to temperature applications.

The package builds on established statistical methodology, including linear
Gaussian state-space modeling, Kalman filtering, and Kalman smoothing. Model
estimation is handled through the `KFAS` package, which provides a general
framework for state-space models in R.

The initial implementation was adapted from the supplementary code provided by
Baba (2024), accompanying Baba et al. (2024), which analyzed sea temperature
trends using a linear Gaussian state-space model. The supplementary code is
publicly available at:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

Compared with that prior implementation, `tempssm` extends the workflow into a
reusable R package interface with input validation, documented S3 methods,
tests, diagnostics, cross-validation utilities, and examples for broader
temperature time-series analysis.

# Model Overview

The model implemented in `tempssm` can be viewed as an extension of the Basic
Structural Time Series Model (BSTSM), a standard state-space formulation that
represents an observed time series using latent trend, seasonal, and irregular
components. Following the temperature time-series application of Baba et al.
(2024), `tempssm` keeps this interpretable decomposition and adds
autoregressive dependence and optional exogenous effects. This makes it useful
for separating long-term temperature change, seasonal cycles, and short-term
departures.

The computational backend is `KFAS`, and the fitted `tempssm` object stores
both filtering and smoothing estimates. Unless otherwise stated, this manual
interprets summary statistics, diagnostics, and plots based on smoothed state
estimates.

The detailed mathematical formulation is separated from this tutorial-style
manual to reduce duplication and improve maintainability. See the package
model specification document for the observation equation, state equations,
seasonal constraint, autoregressive and exogenous components, parameter count,
and estimation procedure. It is available as a PDF in the package repository:

<https://github.com/akihirao/tempssm/blob/main/vignettes/model-specification.pdf>

# How to Use

## Set Environment

Load the following R packages to run the examples below. If any packages are
not installed, please install them as needed.

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
library(forecast)

library(purrr) 
library(tibble)
library(readr)
library(dplyr)
library(ggplot2)

library(patchwork)
```

## Exercise I: Applying a State-Space Model to a Univariate Temperature Time Series

### Objective

In Exercise I, we apply a linear Gaussian state-space model to a univariate
temperature time series. This exercise introduces the basic workflow for model
fitting, component visualization, residual diagnostics, and short-term
prediction without exogenous variables.

### Loading the Sea Surface Temperature (SST) Dataset

A sample sea surface temperature (SST) dataset is included in the package.

- **Dataset**: Monthly sea surface temperature (SST) off Jogashima, Japan  
- **Unit**: °C  
- **Period**: January 1998 to February 2023

This dataset is a simulated monthly SST time series generated from the
state-space model analysis of sea temperature off Jogashima reported by Baba
et al. (2024). It is included as a reproducible example connected to the
motivating study of this package.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(sst_sim) # load a ts object of SST off Jogashima
head(sst_sim)
summary(sst_sim)
```

### Plotting the Monthly SST Time Series

We begin by visualizing the monthly SST time series to examine its
overall structure, including apparent trends, seasonal variability,
and the presence or absence of missing observations.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_sst_sim <- forecast::autoplot(sst_sim) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)") +
  ggtitle("Monthly SST off Jogashima, Japan") +
  theme_classic()

plot(plt_sst_sim)
```

```{r echo = FALSE}
sst_sim_mean <- mean(sst_sim, na.rm = TRUE)
sst_sim_na <- sum(is.na(sst_sim))
```

The overall mean SST is approximately
`r sprintf("%.1f", sst_sim_mean)` °C, and a clear seasonal pattern is
visible. The series contains
`r sst_sim_na` missing observations. The raw time series suggests that
SST may have decreased gradually from the beginning of the record to around
2008 and then increased thereafter. However, interannual variability is also
evident, making the long-term pattern difficult to assess from the raw series
alone.

### Applying a Linear Gaussian State-Space Model

When a `ts` object containing temperature time-series data (here,
`sst_sim`) is passed to the core function `tempssm()`, model construction
and parameter estimation are performed together. The returned S3 object of
class `tempssm` (here, `res_ar1`) stores the filtering and smoothing estimates,
as well as the constructed model and input data. By default, `tempssm()` fits a
first-order autoregressive model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# model with first-order autoregressive component
res_ar1 <- tempssm(sst_sim) # (ar_order=1: default)
summary(res_ar1)
```

From the summary output, confirm that the model has converged
(`Converged: TRUE`). The output also reports statistics such as the number of
parameters (`k`), the log-likelihood, the likelihood type, and the number of
diffuse initial states. The parameter estimates include the observation error
variance (`H`), the process error variance of the long-term trend component
(`Q trend`), the process error variance of the seasonal component
(`Q season`), the process error variance of the autoregressive component, and
the first-order autoregressive coefficient (`AR1`).

The log-likelihood and the associated number of estimated parameters can also
be extracted directly from the fitted `tempssm` object using `logLik()`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
ll <- logLik(res_ar1)
ll
attr(ll, "df") # number of parameters
```

By default, `tempssm()` uses the KFAS marginal likelihood for parameter
estimation. The diffuse likelihood remains available by explicitly setting
`marginal = FALSE`. The selected likelihood type is stored in the fitted
`tempssm` object and is used consistently by downstream methods such as
`logLik()` and `summary()`.

The package intentionally does not compute AIC for `tempssm` objects. This is
because AIC comparisons for state-space models can require care when
likelihood type, diffuse initialization, or latent-state structure differs
across models. To avoid presenting AIC as a default model-selection criterion,
`tempssm` does not provide automatic AIC-based comparison. Instead, this manual
emphasizes a model-assessment workflow based on residual diagnostics and
time-series cross-validation. The log-likelihood and parameter count remain
available through `logLik()` for users who need them for their own
model-assessment workflows.

### Plotting Long-Term Trend, Drift, Seasonal, and Autoregressive Components

We extract the corresponding latent components from the state-space model and
visualize the estimated long-term evolution of temperature levels and their
rate of change (drift). Plotting seasonal variation and autoregressive
dependence as well makes the underlying trend structure easier to examine.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot all components at once
plot(res_ar1)
```

```{r echo = FALSE}
mean_drift_year <- mean(get_drift_ts(res_ar1))
```

The long-term trend in the upper-left panel suggests that SST decreased during
the first part of the study period and then increased after around 2008. In
this example, the average annual rate of SST change over the full period is
approximately `r sprintf("%.3f", mean_drift_year)` °C.

However, this full-period average should be interpreted together with the
time-varying drift component. In the upper-right panel, the drift component is
negative before around 2008 and becomes positive thereafter, corresponding to
the shift from a decreasing to an increasing long-term SST pattern.
The shaded gray areas represent 95% confidence intervals for the estimated
latent states and illustrate the uncertainty associated with each estimated
component.

For scripted workflows, `plot_tempssm_components()` returns the same component
plot as a `ggplot` object; `ggplot2::autoplot()` is also available. To obtain
a ggplot object for an individual component, specify the `component` argument
in `plot_tempssm_components()` as follows.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot each of components at once
plt_level <- plot_tempssm_components(res_ar1, component = c("level"))
plt_drift <- plot_tempssm_components(res_ar1, component = c("drift"))
plt_season <- plot_tempssm_components(res_ar1, component = c("season"))
plt_ar <- plot_tempssm_components(res_ar1, component = c("ar"))
```

### Model Diagnostics

The package provides diagnostic tools for checking whether the fitted model has
left notable structure in the residuals. In particular, residual time-series
plots, residual autocorrelation, residual distributions, and Ljung-Box tests
can be used to assess remaining temporal dependence and departures from the
Gaussian error assumption.

```{r message = FALSE, warning = FALSE, echo = TRUE}
r <- get_tempssm_residuals(res_ar1)
lb_lag <- frequency(res_ar1$temp_data)
forecast::checkresiduals(r, lag = lb_lag, test = "LB")
```

Here, `get_tempssm_residuals()` extracts standardized recursive residuals, and
`forecast::checkresiduals()` displays their time-series plot, ACF plot,
histogram, and Ljung-Box test. The three plots appear in the upper,
lower-left, and lower-right panels, respectively. The lag is set to the
seasonal frequency of the input data; for monthly data, this uses lag 12.
These plots and the test result should be checked for any notable residual
patterns.

In this example, the ACF plot does not show a clear sequence of successive
significant lags or a repeated seasonal pattern. Isolated spikes can occur when
many lags are inspected, so they should be interpreted together with the
Ljung-Box test and the residual time-series plot rather than used as the sole
basis for changing the model.

The package wrapper `plot_tempssm_residual_diagnostics()` is also available
when a single function call for these diagnostic plots is preferred.

For a compact tabular summary of the Ljung-Box test and residual kurtosis, use
`diagnose_residuals()`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag <- diagnose_residuals(res_ar1)
print(diag)
```

The `lb_stat`, `lb_lag`, and `lb_pvalue` columns returned by
`diagnose_residuals()` correspond to the Ljung-Box test statistic, the lag used
in the test, and the corresponding P-value. For monthly time series, the
default Ljung-Box lag is the seasonal frequency, so the default test evaluates
residual autocorrelation up to lag 12. In this example, the Ljung-Box test does
not indicate significant residual autocorrelation up to lag 12.

If residual autocorrelation at longer lags is of concern, the lag can be
specified manually. For example, the following code repeats the Ljung-Box test
up to lags 24 and 36.

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag_lag24 <- diagnose_residuals(res_ar1, lb_lag = 24)
diag_lag36 <- diagnose_residuals(res_ar1, lb_lag = 36)

diag_lags <- rbind(diag, diag_lag24, diag_lag36)
diag_lags$check <- c("lag 12", "lag 24", "lag 36")
diag_lags[, c("check", "lb_stat", "lb_lag", "lb_pvalue", "kurtosis")]
```

These diagnostics should be interpreted together with the residual ACF plot.
A small number of isolated ACF spikes can occur by chance when many lags are
inspected, whereas repeated or periodic spikes may suggest remaining temporal
structure.

### Examining the Autoregressive (AR) Order

In this manual, the default AR(1) specification is used as the working model.
The autoregressive component is included to absorb short-term serial
dependence that is not represented by the latent level, drift, and seasonal
components. A higher AR order should therefore not be treated as an automatic
improvement. Instead, the AR order can be adjusted when residual diagnostics
suggest that notable autocorrelation remains.

For example, if residual autocorrelation remains after fitting the AR(1)
model, users can fit an AR(2) model and repeat the same diagnostics.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Optional sensitivity check with a second-order autoregressive component
res_ar2 <- tempssm(sst_sim, ar_order = 2)
summary(res_ar2)
plot_tempssm_residual_diagnostics(res_ar2)
diag_lag12_ar2 <- diagnose_residuals(res_ar2, lb_lag = 12)
diag_lag24_ar2 <- diagnose_residuals(res_ar2, lb_lag = 24)
diag_lag36_ar2 <- diagnose_residuals(res_ar2, lb_lag = 36)
diag_lags_ar2 <- rbind(diag_lag12_ar2, diag_lag24_ar2, diag_lag36_ar2)
diag_lags_ar2$check <- c("lag 12", "lag 24", "lag 36")
diag_lags_ar2[, c("check", "lb_stat", "lb_lag", "lb_pvalue", "kurtosis")]
```

In a sensitivity check with an AR(2) specification, the residual diagnostics
were very similar to those from the AR(1) model. This suggests that increasing
the AR order does not provide an obvious diagnostic improvement in this
example.

In the present example, the residual diagnostics do not provide strong
evidence that a higher AR order is needed, so we proceed with the AR(1) model
as the baseline specification.

### Estimated Parameters and Latent-State Components

The fitted `tempssm` object contains the smoothed state estimates returned by
`KFAS`. The matrix `res_ar1$kfs$alphahat` stores these latent-state estimates,
including the level, drift, seasonal states, and autoregressive state.
Inspecting this matrix is useful for understanding the internal state
representation.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res_ar1$kfs$alphahat
head(alpha_hat)
```

For routine use, helper functions provide a simpler way to extract individual
components as `ts` objects with the original time index. The level component
represents the estimated long-term temperature level, while the drift
component represents its rate of change per year.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimate of level component
level_ts <- get_level_ts(res_ar1)

# Smoothing estimate of drift component
drift_ts <- get_drift_ts(res_ar1)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts) 
print(mean_drift_year)
```
 
The average annual rate of SST change was estimated to be
`r sprintf("%.4f", mean_drift_year)` °C over the full period. Because the
estimated drift changes sign over time, this value should be interpreted as a
model-based summary of the whole study period rather than as a constant
warming rate.

### Short-Term Prediction

A fitted `tempssm` object can also be passed to `predict()` to obtain
short-term predictions. By default, `predict(res)` returns a one-step-ahead
prediction beyond the end of the observed series. This is useful for visual
checks of how the fitted model extrapolates the estimated level, seasonal, and
autoregressive components.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_1 <- predict(res_ar1)
pred_1
```

Predictions for multiple future time points can be requested by setting the
`n.ahead` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12 <- predict(res_ar1, n.ahead = 12)
pred_12
```

Prediction uncertainty can also be returned by setting the `interval`
argument. With `interval = "confidence"`, the returned lower and upper bounds
describe uncertainty in the predicted mean response. With
`interval = "prediction"`, the bounds describe uncertainty in a future
observation and therefore also include observation error; prediction intervals
are usually wider than confidence intervals. The confidence level is controlled
by the `level` argument and defaults to 0.95.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12_pi <- predict(
  res_ar1,
  n.ahead = 12,
  interval = "prediction",
  level = 0.95
)

pred_12_pi
```

These predictions should be interpreted as model-based extrapolations rather
than definitive forecasts. Uncertainty generally increases as the prediction
horizon becomes longer, and long-horizon predictions can be sensitive to model
assumptions about trend, seasonality, and autoregressive dependence.


## Exercise II: Model With Exogenous Variables

### Objective

We extend the state-space modeling framework to examine the effect of an
exogenous factor on temperature variation. Specifically, we examine simulated
Kuroshio path information as exogenous variables for the simulated SST series
introduced in Exercise I.

### Loading Simulated Kuroshio Exogenous Variables

Two simulated exogenous-variable datasets are included in the package.

- **`kuroshio_a_sim`**: monthly binary indicator of the Kuroshio A-type path  
- **`distance_sim`**: monthly offshore-distance index of the Kuroshio path  
- **Period**: January 1998 to February 2023

Both datasets are provided as univariate `ts` objects with `frequency = 12`.
They are simulated monthly datasets based on the Kuroshio-related variables
used in Baba et al. (2024). The original data used in that study had finer
time resolution for some variables, but the simulated datasets in this package
are represented as monthly time series for use with `tempssm` examples.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(kuroshio_a_sim)
data(distance_sim)

head(kuroshio_a_sim)
head(distance_sim)

summary(kuroshio_a_sim)
summary(distance_sim)
```

### Checking Time-Series Alignment

For state-space modeling with exogenous variables, the response and exogenous
time series must share a common and aligned time index. In this example,
`sst_sim`, `kuroshio_a_sim`, and `distance_sim` already have the same monthly
time span and frequency.

```{r message = FALSE, warning = FALSE, echo = TRUE}
series_info <- tibble::tibble(
  series = c("sst_sim", "kuroshio_a_sim", "distance_sim"),
  start = c(
    paste(start(sst_sim), collapse = "-"),
    paste(start(kuroshio_a_sim), collapse = "-"),
    paste(start(distance_sim), collapse = "-")
  ),
  end = c(
    paste(end(sst_sim), collapse = "-"),
    paste(end(kuroshio_a_sim), collapse = "-"),
    paste(end(distance_sim), collapse = "-")
  ),
  frequency = c(
    frequency(sst_sim),
    frequency(kuroshio_a_sim),
    frequency(distance_sim)
  ),
  missing = c(
    sum(is.na(sst_sim)),
    sum(is.na(kuroshio_a_sim)),
    sum(is.na(distance_sim))
  )
)

series_info
```

If user-supplied time series have different start or end points,
`trim_ts_overlap()` can be used to restrict the response and exogenous series
to their shared overlapping period before fitting the model.

### Plotting the Response and Exogenous Variables

We visualize the simulated SST series together with the two simulated Kuroshio
variables to inspect the overall structure of the dataset.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_sst_exo <- forecast::autoplot(sst_sim) +
  labs(y = expression(Temp.~(degree*C)), x = "Time (year)") +
  ggtitle("Simulated monthly SST") +
  theme_classic()

plt_kuroshio_a <- forecast::autoplot(kuroshio_a_sim) +
  labs(x = "Time (year)", y = "A-type indicator") +
  ggtitle("Simulated Kuroshio A-type path") +
  scale_y_continuous(breaks = c(0, 1)) +
  theme_classic()

plt_distance <- forecast::autoplot(distance_sim) +
  labs(x = "Time (year)", y = "Offshore distance") +
  ggtitle("Simulated Kuroshio offshore distance") +
  theme_classic()

plt_sst_exo + plt_kuroshio_a + plt_distance +
  patchwork::plot_layout(ncol = 1)
```

The simulated exogenous-variable series used in this exercise contain no
missing values. When using your own exogenous-variable data, make sure that
the exogenous variables do not contain missing values before fitting the
model. `tempssm()` does not allow missing values in exogenous variables, and
model fitting will stop with an error if such data are supplied. If necessary,
apply appropriate preprocessing, such as imputation, before the analysis.

In contrast, missing values in the dependent temperature time series are
allowed. In the state-space model, these missing values are treated as
unobserved responses, and model estimation proceeds through Kalman filtering
and smoothing.

### Applying a Model Without an Exogenous Variable

The baseline model without exogenous variables was already fitted in
Exercise I. Because Exercise II uses the same response series, `sst_sim`, we
reuse that fitted object as the reference model instead of refitting the same
model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_without <- res_ar1
summary(res_without)
```

This avoids repeating the same estimation step and makes the comparison focus
on the additional information introduced by the simulated Kuroshio variables.

### Combining Exogenous Variables into a Multivariate `ts` Object

The `exo_data` argument of `tempssm()` can accept either a univariate `ts`
object or a multivariate `ts` object. Because this exercise uses two exogenous
variables, we combine `kuroshio_a_sim` and `distance_sim` into a single
multivariate `ts` object before fitting the model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
exo_kuroshio <- cbind(kuroshio_a_sim, distance_sim)
colnames(exo_kuroshio) <- c("kuroshio_a", "distance")

head(exo_kuroshio)
start(exo_kuroshio)
end(exo_kuroshio)
frequency(exo_kuroshio)
colnames(exo_kuroshio)
sum(is.na(exo_kuroshio))
```

The resulting object has class `mts`, which is the multivariate form of base
R's `ts` class. Each column is treated as a separate exogenous variable in the
state-space model. The column names are used as variable labels in summaries
and coefficient tables, so assigning informative names before model fitting is
helpful.

### Applying a Model With Exogenous Variables

We next construct a model that uses the two simulated Kuroshio variables as
exogenous variables. In the `tempssm()` function, the multivariate object
`exo_kuroshio` is supplied to the exogenous-variable argument (`exo_data`),
and the simulated SST series (`sst_sim`) is supplied to the
dependent-variable argument (`temp_data`).

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_with <- tempssm(
  temp_data = sst_sim,
  exo_data = exo_kuroshio,
  ar_order = 1
)
summary(res_with)

plot_tempssm_residual_diagnostics(res_with)
diag_res_with <- diagnose_residuals(res_with)
print(diag_res_with)
```

```{r echo = FALSE}
exo_coef <- get_exo_coef(res_with)
```

The residual diagnostics for the model with the simulated Kuroshio variables
also do not indicate significant residual autocorrelation up to lag 12. Thus,
for both models considered here, the AR(1) specification provides an adequate
working autoregressive structure according to these residual checks.

We then examine the estimated coefficients for the exogenous variables.

```{r message = FALSE, warning = FALSE, echo = TRUE}
exo_coef
```

The coefficient table summarizes the estimated associations between SST and
each simulated Kuroshio variable after accounting for the latent long-term
trend, seasonal cycle, and autoregressive dependence. Because these are
simulated example datasets, the estimates should be interpreted as an
illustration of how to fit and inspect a model with multiple exogenous
variables, rather than as new scientific evidence about the Kuroshio path.

For models with exogenous variables, future values of the exogenous variables
are needed when calling `predict()`. If such values are available, they can be
passed through `new_exo_data`. For a model with multiple exogenous variables,
`new_exo_data` should contain the same variables, in the same order, as the
object supplied to `exo_data`. For a simple one-step-ahead visual check,
`exo_strategy = "last"` carries the final observed row of all exogenous
variables forward. This is a persistence assumption for the exogenous
variables, not a separate forecast model for the Kuroshio path.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_with_last_exo <- predict(res_with, exo_strategy = "last")
pred_with_last_exo
```

The next section evaluates whether including these simulated Kuroshio
variables improves out-of-sample predictive performance under a conditional
forecasting setting.

### Time-Series Cross-Validation (tsCV)

Time-series cross-validation (tsCV) evaluates model performance based on
out-of-sample prediction errors while respecting the temporal ordering of the
data. It therefore provides an additional and independent perspective on model
adequacy.

In time-series cross-validation, the model is repeatedly fitted to expanding or
sliding training windows, and predictive performance is evaluated on subsequent
observations. Unlike random cross-validation, this procedure prevents
information leakage from the future to the past and is therefore suitable for
model validation with temporal data.

Here, by comparing cross-validation metrics for models with and without the
simulated Kuroshio variables, we assess whether the additional exogenous
information is reflected in out-of-sample predictive performance.

For the model with the simulated Kuroshio variables, the test-period values of
those variables are supplied to the forecast step as exogenous variables.
Therefore, this tsCV setting evaluates conditional forecasts, or hindcasts,
under the assumption that the Kuroshio-variable values during each test period
are known. This is appropriate when the aim is to assess the explanatory or
reconstructive value of the exogenous variables.

For real-time future forecasting, however, future exogenous-variable values are
usually unknown. In that setting, users must either provide future scenarios,
forecast the exogenous variables with separate models, or use a simplified
assumption such as carrying the last observed value forward. The results below
should therefore be interpreted as evidence for improved conditional
predictive performance, not as a direct guarantee of operational forecast
improvement.


```{r message = FALSE, warning = FALSE, echo = TRUE}

# (Optional) Load packages for parallel processing.
# These are only required if you want to enable parallel execution:
# - future: controls how parallel workers are launched
# - future.apply: provides parallelized apply functions
# - progressr: optional progress bar support
library(future.apply)
library(future)
library(progressr)

if (interactive()) {
  future::plan(multisession)
} else {
  future::plan(sequential)
}


## Generate a list of training and test datasets with their indices
# Procedure for constructing year-based time-series cross-validation folds:
# 
# First training data: January 1998-December 2010;
# First test data: January 2011-December 2011
# 
# Second training data: January 1999-December 2011;
# Second test data: January 2012-December 2012
# ...
# 
# Twelfth training data: January 2009-December 2021;
# Twelfth test data: January 2022-December 2022
#
# These folds are automatically generated by the ts_train_test_split() function.
# The final two observations in 2023 are not used because allow_partial = FALSE
# keeps every test set as a complete 12-month period.

# Generate training and test dataset for model without exogenous variables
folds_without <- ts_train_test_split(
  temp_data = sst_sim,
  exo_data = NULL,
  initial = 156, # 156 monthly observations from Jan 1998 to Dec 2010
  horizon = 12, # forecast 12 monthly observations
  step = 12, # move the fixed-width window in one-year steps
  fixed_window = TRUE,
  allow_partial = FALSE
  )

# Generate training and test dataset for model with Kuroshio variables
folds_with <- ts_train_test_split(
  temp_data = sst_sim,
  exo_data = exo_kuroshio,
  initial = 156, # 156 monthly observations from Jan 1998 to Dec 2010
  horizon = 12, # forecast 12 monthly observations
  step = 12, # move the fixed-width window in one-year steps
  fixed_window = TRUE,
  allow_partial = FALSE
  )

# Check the first fold
start(folds_without[[1]]$train_ts)
end(folds_without[[1]]$train_ts)
start(folds_without[[1]]$test_ts)
end(folds_without[[1]]$test_ts)


# *****************************************
# Executing time-series cross validation

#-------------------------------------------------- 
# Model without exogenous variables

# Single processing
# cv_without_results <- ts_cv_run(
#   folds_without,
#   ar_order = 1,
#   use_season = TRUE
# )

# Parallel processing
cv_without_results <- future.apply::future_lapply(
  folds_without,
  ts_cv_run_fold,
  ar_order = 1,
  use_season = TRUE,
  future.seed = TRUE
)

# Computing assessment indexes
metrics_without <- lapply(cv_without_results, compute_cv_metrics)

# tidy summary
cv_without_tbl <- ts_cv_collect(cv_without_results, metrics_without) %>%
  mutate(Model="Without")


#-------------------------------------------------- 
# Model with simulated Kuroshio exogenous variables

# Single processing
# cv_with_results <- ts_cv_run(folds_with, ar_order = 1, use_season = TRUE)

# Parallel processing
cv_with_results <- future.apply::future_lapply(
  folds_with,
  ts_cv_run_fold,
  ar_order = 1,
  use_season = TRUE,
  future.seed = TRUE
)

# Computing assessment indexes
metrics_with <- lapply(cv_with_results, compute_cv_metrics)

# tidy summary
cv_with_tbl <- ts_cv_collect(cv_with_results, metrics_with) %>%
  mutate(Model="With")


cv_tbl <- bind_rows(cv_without_tbl, cv_with_tbl)

cv_comparison <- compare_ts_cv(
  list(
    Without = cv_without_tbl,
    With = cv_with_tbl
  )
)

cv_comparison %>% knitr::kable()


plt_MAE <- ggplot(data=cv_tbl,
                  aes(x=Model,y=MAE)) +
  geom_boxplot()

plt_MASE_naive <- ggplot(data=cv_tbl,
                   aes(x=Model,y=MASE_naive)) +
  geom_boxplot()

plt_MASE_seasonal <- ggplot(data=cv_tbl,
                   aes(x=Model,y=MASE_seasonal)) +
  geom_boxplot()


plt_tsCV <- plt_MAE + plt_MASE_naive + plt_MASE_seasonal +
  patchwork::plot_layout(nrow = 1)

plot(plt_tsCV)

```

```{r echo = FALSE}
monthly_mean_sst <- tapply(
  as.numeric(sst_sim),
  cycle(sst_sim),
  mean,
  na.rm = TRUE
)
seasonal_amplitude <- diff(range(monthly_mean_sst, na.rm = TRUE))
mae_without <- cv_comparison$mean_MAE[cv_comparison$model == "Without"]
mae_with <- cv_comparison$mean_MAE[cv_comparison$model == "With"]
mae_relative_to_season <- 100 * mae_with / seasonal_amplitude
mae_improvement <- 100 * (mae_without - mae_with) / mae_without
n_folds <- unique(cv_comparison$n_folds)
n_folds <- if (length(n_folds) == 1L) n_folds else min(n_folds)
```

The tsCV results show that the model with the exogenous variable performs
better than the model without it. The comparison table produced by
`compare_ts_cv()` summarizes the number of folds, convergence rates, and mean
prediction-error metrics for each model, while the boxplots show how the
fold-level errors vary across validation periods. In this tutorial, the number
of tsCV iterations is set to `r n_folds` to reduce execution time. In actual
validation, please use a sufficient number of iterations.

The mean MAE of the model with the exogenous variable is approximately
`r sprintf("%.2f", mae_with)` °C. Relative to the seasonal amplitude based on
monthly mean SST (about `r sprintf("%.1f", seasonal_amplitude)` °C), this
corresponds to roughly `r sprintf("%.1f", mae_relative_to_season)`%. Thus, the
prediction error is small compared with the dominant seasonal scale of
variation in this data set, and may be practically useful for short-term marine
monitoring.

Although the number of validation folds is limited to `r n_folds`, the model
with the exogenous variable reduces the mean MAE by approximately
`r sprintf("%.1f", mae_improvement)`% compared with the model without the
exogenous variable. Similar improvements are also seen in the MASE metrics.
This interpretation should be kept conditional on the tsCV setting used here,
in which the test-period values of the simulated Kuroshio variables are
supplied as exogenous variables.

Taken together, the results illustrate how exogenous-variable models can be
evaluated using both coefficient estimates and time-series cross-validation.
Because the datasets in this exercise are simulated, the emphasis is on the
workflow rather than on drawing new scientific conclusions about Kuroshio
variability.

These complementary lines of evidence can help users decide whether an
exogenous-variable model is useful for their own data, while keeping the
forecasting assumptions for future exogenous variables explicit.

This exercise used two exogenous variables supplied as multivariate
time-series data. Users interested in broader applications can extend the same
workflow to other environmental, climatic, or ecological covariates. In doing
so, model structure should be considered carefully for the target data, while
checking model diagnostics and out-of-sample predictive performance.


### Comparing Estimation Patterns Between Models

We compare how the estimated patterns of the long-term trend and its rate of
change differ between the models with and without the simulated Kuroshio
variables.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_level_without_ts <- autoplot(res_without,
                                 component=c("level")
                                 ) +
  labs(title="Model without exogenous variables") +
  theme_classic()

plt_drift_without_ts <- autoplot(res_without, 
                             component=c("drift")
                             ) + 
  labs(title="Model without exogenous variables") +
  theme_classic()

plt_level_drift_without_ts <- plt_level_without_ts + 
  plt_drift_without_ts + 
  patchwork::plot_layout(ncol=1)

plot(plt_level_drift_without_ts)


plt_level_with_ts <- autoplot(res_with,
                          component=c("level")
                          )+ 
  labs(title="Model with Kuroshio variables") +
  theme_classic()

plt_drift_with_ts <- autoplot(res_with,
                          component=c("drift")
                          ) + 
  labs(title="Model with Kuroshio variables") +
  theme_classic()

plt_level_drift_with_ts <- plt_level_with_ts + 
  plt_drift_with_ts + 
  patchwork::plot_layout(ncol=1)

plot(plt_level_drift_with_ts)

#　Smoothing estimate of drift component
mean_drift_year_without <- get_drift_ts(res_without) %>%
  mean()
print(mean_drift_year_without)

mean_drift_year_with <- get_drift_ts(res_with) %>%
  mean()
print(mean_drift_year_with)

```
The gray areas in the plots above show 95% confidence intervals.

In the model without an exogenous variable, the plots of the long-term trend
and its rate of change show a wavelike pattern. In contrast, in the model with
the exogenous variable, the long-term trend and its rate of change show a more
stable pattern. This suggests that the wavelike pattern reflects the influence
of variation represented by the simulated Kuroshio variables.

The estimated annual rate of SST change over the observation period is
`r sprintf("%.4f", mean_drift_year_without)` in the model without the
exogenous variable and `r sprintf("%.4f", mean_drift_year_with)` in the model
with the exogenous variable.

In this simulated example, the model incorporating the exogenous variables
provides an alternative decomposition of the SST series after accounting for
the additional covariates. The comparison illustrates that estimated long-term
components can change when external drivers are included, which is one reason
to inspect both model diagnostics and predictive performance before selecting a
working model.


# Using Your Own Data

## Quick Example

`my_ts` is a user-supplied `ts` object containing temperature time-series data.
It can be passed directly to `tempssm()`.

```{r, eval = FALSE}
## example of ts object (not run)
## my_ts <- ts(rnorm(100), frequency = 12)
res <- tempssm(my_ts)
```

## Preparing Your Own Dataset

### CSV Format
Prepare a CSV file of monthly temperature time series with the following
columns:
- Year
- Month
- Temp

Example:
```text
Year,Month,Temp
2010,8,13.6
2010,9,6.8
2010,10,NA
2010,11,-1.4
...
```

* Use NA for missing temperature values, and always keep the corresponding Year
  and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.


### Converting CSV to a Time Series

An example CSV file included in this package is available in inst/extdata. The
example dataset contains monthly air temperature observations from Mt. Akadake,
Hokkaido, Japan (1,840 m). The original data are available from the Monitoring
Sites 1000 Project of the Ministry of the Environment of Japan (KOZ01.zip,
https://www.biodic.go.jp/moni1000/findings/data/index.html).

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")
akadake_temp <- tempssm::read_monthly_temp_ts(path)
head(akadake_temp)
```

The `read_monthly_temp_ts()` function reads this type of CSV file and converts
it into an R `ts` object for use with `tempssm()`. If finer control is needed,
you can also construct a `ts` object manually; see the `stats::ts`
documentation for details:
<https://search.r-project.org/R/refmans/stats/html/ts.html>.



# References

The statistical modeling framework implemented in `tempssm` is based on the
methodology described in Baba et al. (2024), and the accompanying supplementary
materials and code repository served as the initial source for development of
this package.

Baba, S., Ishii, H., and Yoshiyama, T. (2024).
Estimating sea temperature trends using a linear Gaussian state-space model
in Jogashima, Kanagawa, Japan.
*Bulletin of the Japanese Society of Fisheries Oceanography*,
88(3), 190–199.
(In Japanese with an English abstract.)
https://doi.org/10.34423/jsfo.88.3_190

Baba, S. (2024).
Supplementary code and test data for estimating sea temperature trends using a
linear Gaussian state-space model.
GitHub repository:
https://github.com/logics-of-blue/sea-temperature-trend-jogashima

Helske, J. (2017).
KFAS: Exponential Family State Space Models in R.
*Journal of Statistical Software*, 78(10), 1–39.
https://doi.org/10.18637/jss.v078.i10

# Appendix: Utility Functions

The following utility functions are provided to support data preparation and
exploratory analysis.


## 1. `read_monthly_temp_ts()`

The `read_monthly_temp_ts()` function converts monthly temperature data stored
in a CSV file into an R `ts` object. By enforcing a simple and consistent data
format, it is intended to make externally prepared time-series data easier to
import into tempssm.

### Example

For monthly temperature data, prepare a CSV file with a header row. By default,
the column names must be `Year`, `Month`, and `Temp`, where `Temp` represents
the observed temperature value.

```text
Year,Month,Temp
2001,1,10.4
2001,2,8.2
2001,3,NA
2001,4,13.6
2001,5,16.1
...
```

* Use NA for missing temperature values, and always keep the corresponding Year
  and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.

The following example uses the sample CSV file included in the package. It
contains monthly air temperature observations from Mt. Akadake, Hokkaido,
Japan.

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")

# Read the CSV file and convert it to a monthly ts object
temp_ts <- read_monthly_temp_ts(path)
head(temp_ts)
```

## 2. `convert_monthly_df_to_ts()`

The `convert_monthly_df_to_ts()` function converts a data frame containing
monthly temperature data into an R `ts` object. It is designed to support
workflows in which temperature data are first imported or prepared as a data
frame and then used for time-series analysis.

The input data frame is expected to contain at least two columns: a date column
(`Date`) and a temperature column (`Temp`). For the date column (`Date`), assign
an arbitrary day of the month, such as the first day, so that the dates
correspond to monthly aggregated temperature values.

### Example

```{r message = FALSE, warning = FALSE}
# Create a data frame of monthly temperature data
df <- data.frame(
  Date = as.Date(c(
    "2001-01-01",
    "2001-02-01",
    "2001-03-01",
    "2001-04-01",
    "2001-05-01")
    ),
  Temp = c(10.4, 8.2, NA, 13.6, 16.1)
  )

# Convert to a monthly ts object
temp_ts <- convert_monthly_df_to_ts(df)
head(temp_ts)
```

## 3. `get_jma_sst_ts()`

The `get_jma_sst_ts()` function downloads daily mean sea surface temperature
(SST) data for Japanese coastal waters provided by the Japan Meteorological
Agency (JMA). It aggregates the daily values into monthly means and returns the
resulting time series as an object of class `ts`.

### Example

The following example demonstrates how to download SST data for the southern
coastal waters of Ibaraki Prefecture, Japan.

The argument `sea_area_id` specifies the numeric identifier of a sea area
defined by JMA. For example, `138` corresponds to the coastal waters off
southern Ibaraki. A list of available sea area IDs and their corresponding
regions is provided by JMA at:

<https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html>

```{r message = FALSE, warning = FALSE, echo = TRUE}
sst_138_ts <- get_jma_sst_ts(sea_area_id = 138)
head(sst_138_ts)
```

## 4. `compute_temp_anomaly()`

The `compute_temp_anomaly()` function computes anomalies from a time series
provided as an R `ts` object. Anomalies are calculated by subtracting the
corresponding long-term mean for each periodic unit, such as month, from each
observation. This transformation is useful for examining interannual
variability after removing periodic effects.

The reference period used to compute anomalies can be specified via a function
argument. By default, the climatology is computed over the entire available
time series (see `?compute_temp_anomaly`).

This transformation is useful for exploratory analysis and modeling
applications that focus on departures from typical seasonal conditions.

### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
data(niigata_sst)
niigata_sst_anomaly <- compute_temp_anomaly(niigata_sst)
plt_niigata_sst_anomaly <- forecast::autoplot(niigata_sst_anomaly) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)")
plot(plt_niigata_sst_anomaly) 
```

## 5. `compute_monthly_climatology()`

The `compute_monthly_climatology()` function computes the climatological mean
seasonal cycle from a `ts` object by averaging each seasonal value across
years. It is primarily intended for exploratory analysis and visualization of
seasonal structure in temperature time series.

### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(niigata_sst)
monthly_seasonal_cycle_niigata_sst <- compute_monthly_climatology(niigata_sst) 
summary(monthly_seasonal_cycle_niigata_sst)

plt_monthly_seasonal_cycle_niigata_sst <- ggplot(
  data = monthly_seasonal_cycle_niigata_sst,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temp.~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  ) +
  theme_classic()


plot(plt_monthly_seasonal_cycle_niigata_sst)
```
