---
title: "tempssm Reference Manual"
subtitle: "Version `r packageVersion('tempssm')`"
author: "Akira S. Hirao, Shinya Baba, and Momoko Ichinokawa"
date: "`r Sys.Date()`"
output:
#  pdf_document:
#    latex_engine: lualatex
#    toc: true
  rmarkdown::html_vignette: default
#  md_document:
#    variant: gfm+tex_math_dollars
vignette: >
  %\VignetteIndexEntry{tempssm Reference Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Summary

`tempssm` is an R package for state-space modeling of environmental
temperature time series, including air and water temperature observations. It
provides a practical framework for assessing how long-term trend, seasonal
variation, autoregressive dependence, and optional exogenous effects contribute
to observed temporal variation. The package facilitates the application of
linear Gaussian state-space models estimated by Kalman filtering and smoothing,
using the `KFAS` package as the computational backend (Helske, 2017).

# Key Features

- Fits linear Gaussian state-space models to environmental temperature time
  series.
- Represents temperature dynamics using interpretable latent components:
  long-term trend, seasonal variation, autoregressive dependence, and optional
  exogenous effects.
- Supports arbitrary seasonal frequencies, while the current examples
  and validation focus primarily on monthly temperature data.
- Allows users to specify an arbitrary order of the autoregressive component
  (default: AR(1)).
- Includes time-series cross-validation tools for model evaluation.

# Input Data Format

R `ts` objects are the primary input format for `tempssm`. The temperature
series passed to `tempssm()` should be supplied as a univariate `ts` object,
and optional exogenous variables can be supplied as univariate or multivariate
`ts` objects. The `ts` class is base R's standard format for regularly spaced
time series (see the `stats::ts` documentation:
<https://search.r-project.org/R/refmans/stats/html/ts.html>). The package also
provides utility functions for converting common tabular and observational
data formats into `ts` objects before model fitting.

The seasonal cycle used by `tempssm()` is taken from the `frequency` attribute
of the input `ts` object. For example, `frequency = 12` represents monthly
data.

# Prior Art and Scope

`tempssm` provides a domain-focused workflow for analyzing temperature time
series with linear Gaussian state-space models. It brings together model
construction, component extraction, uncertainty summaries, residual
diagnostics, visualization, and time-series cross-validation in a single R
package interface tailored to temperature applications.

The package builds on established statistical methodology, including linear
Gaussian state-space modeling, Kalman filtering, and Kalman smoothing. Model
estimation is handled through the `KFAS` package, which provides a general
framework for state-space models in R.

The initial implementation was adapted from the supplementary code provided by
Baba (2024), accompanying Baba et al. (2024), which analyzed sea temperature
trends using a linear Gaussian state-space model. The supplementary code is
publicly available at:

<https://github.com/logics-of-blue/sea-temperature-trend-jogashima>

Compared with that prior implementation, `tempssm` extends the workflow into a
reusable R package interface with input validation, documented S3 methods,
tests, diagnostics, cross-validation utilities, and examples for broader
temperature time-series analysis.

# Model Overview

The model implemented in `tempssm` can be viewed as an extension of the Basic
Structural Time Series Model (BSTSM), a standard state-space formulation that
represents an observed time series using latent trend, seasonal, and irregular
components. Following the temperature time-series application of Baba et al.
(2024), `tempssm` keeps this interpretable decomposition and adds
autoregressive dependence and optional exogenous effects. This makes it useful
for separating long-term temperature change, seasonal cycles, and short-term
departures.

The model is estimated with `KFAS`, and `tempssm()` returns both filtering and
smoothing estimates. Unless otherwise stated, the summaries, diagnostics, and
plots in this vignette use smoothed state estimates.

The full mathematical specification, including the observation equation, state
decomposition, seasonal constraint, autoregressive component, exogenous
component, parameter count, and estimation procedure, is described in the
separate model specification document, available as a PDF in the package
repository:

<https://github.com/akihirao/tempssm/blob/main/vignettes/model-specification.pdf>

# Tutorial Workflow

This tutorial demonstrates a typical workflow for applying `tempssm` to
monthly environmental temperature time series. Exercise I uses the simulated
sea surface temperature (SST) dataset `sst_sim` to fit a baseline state-space 
model without exogenous variables, inspect latent components, check residual 
diagnostics, and obtain short-term predictions. Exercise II extends the same 
workflow by adding simulated marine-environment variables and evaluating 
whether they improve model interpretation and conditional predictive 
performance using diagnostics and time-series cross-validation.

The aim is not to draw new scientific conclusions from the simulated datasets,
but to show how the package can be used to structure model fitting,
diagnostics, prediction, and model comparison workflows.

## Setup

Load the following R packages to run the examples below. If any packages are
not installed, please install them as needed.

```{r message = FALSE, warning = FALSE, echo = TRUE}
## Set libraries
library(tempssm)
library(forecast)

library(purrr) 
library(tibble)
library(readr)
library(dplyr)
library(ggplot2)

library(patchwork)
```

## Exercise I: Baseline Model for a Univariate Temperature Series

Exercise I introduces the basic workflow for fitting a univariate model,
visualizing latent components, checking residual diagnostics, and making
short-term predictions without exogenous variables.

### Load the Simulated SST Dataset

The package includes `sst_sim`, a simulated monthly sea surface temperature
(SST) dataset.

- **Dataset**: Simulated monthly SST off Jogashima, Japan  
- **Unit**: °C
- **Period**: January 1998 to February 2023

The dataset was generated from a state-space model analysis of sea temperature
off Jogashima reported by Baba et al. (2024). It is used here as a
reproducible response series for demonstrating the basic `tempssm` workflow.
Exercise II combines the same simulated SST series with simulated
marine-environment variables.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(sst_sim) # load a ts object of SST off Jogashima
head(sst_sim)
summary(sst_sim)
```

### Plot the Monthly SST Series

We begin by visualizing the monthly SST time series to examine its
overall structure, including apparent trends, seasonal variability,
and the presence or absence of missing observations.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_sst_sim <- forecast::autoplot(sst_sim) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)") +
  ggtitle("Monthly SST off Jogashima, Japan") +
  theme_classic()

plot(plt_sst_sim)
```

```{r echo = FALSE}
sst_sim_mean <- mean(sst_sim, na.rm = TRUE)
sst_sim_na <- sum(is.na(sst_sim))
```

The overall mean SST is approximately
`r sprintf("%.1f", sst_sim_mean)` °C, and a clear seasonal pattern is
visible. The raw time series suggests that SST may have decreased 
gradually from the beginning of the record to around 2008 and then 
increased thereafter. However, interannual variability is also 
evident, making the long-term pattern difficult to assess from 
the raw series alone.

### Fit the Baseline State-Space Model

When a `ts` object containing temperature time-series data (here,
`sst_sim`) is passed to the core function `tempssm()`, model construction
and parameter estimation are performed together. The returned S3 object of
class `tempssm` (here, `res_base`) stores the filtering and smoothing
estimates, as well as the constructed model and input data.

The simulated SST series used in this manual was generated under an AR(2)
setting based on the original analysis of Baba et al. (2024). We therefore use
an AR(2) specification as the baseline model in this detailed example. This is
not a general recommendation that AR(2) should always be preferred. When such
prior information is unavailable, a practical workflow is to start from the
default AR(1) specification and then examine AR(2) or higher-order alternatives
through residual diagnostics and sensitivity checks.

Here, AR means autoregressive. In `tempssm`, the autoregressive component
captures short-term serial dependence that remains after accounting for the
long-term level, drift, and seasonal components. An AR(2) specification means
that this short-term component is modeled using dependence on the previous two
time points.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# baseline model with a second-order autoregressive component
res_base <- tempssm(sst_sim, ar_order = 2)
summary(res_base)
```

From the summary output, confirm that the model has converged
(`Converged: TRUE`). The output also reports the number of parameters (`k`),
the log-likelihood, the likelihood type, and the number of diffuse initial
states.

The parameter estimates correspond to the error variances and autoregressive
coefficients in the fitted state-space model. `H` is the observation error
variance, representing variability in the observed temperature series that is
not explained by the latent states. The `Q` terms are process error variances
for the latent components: `Q trend` controls stochastic variation in the
long-term trend, `Q season` controls variation in the seasonal component, and
the autoregressive process variance controls variation in the short-term
autoregressive component. The `AR` coefficients represent serial dependence in
that short-term component.

The log-likelihood and parameter count can also be extracted directly with
`logLik()`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
ll <- logLik(res_base)
ll
attr(ll, "df") # number of parameters
```

By default, `tempssm()` uses the KFAS marginal likelihood. The diffuse
likelihood remains available with `marginal = FALSE`, and the selected
likelihood type is retained by `logLik()` and `summary()`.

The package intentionally does not compute AIC for `tempssm` objects because
AIC comparisons for state-space models can require care when likelihood type,
diffuse initialization, or latent-state structure differs across models. This
manual instead emphasizes residual diagnostics and time-series
cross-validation. The log-likelihood and parameter count remain available for
users who need them in their own workflows.

### Plot Latent Components

The component plot visualizes the estimated long-term level, drift, seasonal
variation, and autoregressive dependence.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# plot all components at once
plot(res_base)
```

```{r echo = FALSE}
# the average annual rate of SST change over the full period
mean_drift_year <- mean(get_drift_ts(res_base))
mean_drift_year
```

The long-term trend in the upper-left panel suggests that SST decreased during
the first part of the study period and then increased after around 2008. In
this example, the average annual rate of SST change over the full period is
approximately `r sprintf("%.3f", mean_drift_year)` °C.

This full-period average should be interpreted together with the time-varying
drift component. In the upper-right panel, the drift is negative before around
2008 and becomes positive thereafter, corresponding to the shift from a
decreasing to an increasing long-term SST pattern. The shaded gray areas show
95% confidence intervals for the estimated latent states.

For scripted workflows, `plot_tempssm_components()` returns the same component
plot as a `ggplot` object. Individual components can be selected with the
`component` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# extract individual component plots
plt_level <- plot_tempssm_components(res_base, component = "level")
plt_drift <- plot_tempssm_components(res_base, component = "drift")
plt_season <- plot_tempssm_components(res_base, component = "season")
plt_ar <- plot_tempssm_components(res_base, component = "ar")
```

### Model Diagnostics

Residual diagnostics help check whether notable structure remains after model
fitting. Here we inspect the residual series, residual autocorrelation,
residual distribution, and a Ljung-Box test.

```{r message = FALSE, warning = FALSE, echo = TRUE}
resid <- get_tempssm_residuals(res_base)
plot_tempssm_residuals(resid)
```

Here, `get_tempssm_residuals()` extracts standardized recursive residuals, and
`plot_tempssm_residuals()` displays the residual series, ACF plot, and
histogram. The residual series appears in the upper panel, the ACF plot in the
lower-left panel, and the residual frequency distribution in the lower-right
panel.

In this manual, the recommended residual-diagnostic workflow first extracts
residuals as a time-preserving `ts` object and then passes that residual object
explicitly to diagnostic functions. This makes it easier to see which time
points are included in the diagnostic assessment and whether any residuals are
unavailable.

By default, `get_tempssm_residuals()` returns a `ts` object that preserves the
time index of the input series. Missing or non-finite residuals are represented
as `NA`. For calculations that require a complete numeric vector, use
`get_tempssm_residuals(res, keep_time = FALSE)`.

When standardized recursive residuals are extracted as a time-preserving `ts`
object, some initial residuals may be returned as `NA`. In state-space models
with diffuse initialization, standardized recursive residuals are not always
available during the diffuse phase. These `NA` values therefore do not
necessarily indicate that the fitted model failed at those time points. Rather,
they indicate that comparable standardized recursive residuals are not defined
until the diffuse effects have been resolved.

For seasonal models, this can occur even when no exogenous variables are used.
The seasonal states are initialized diffusely, and a full seasonal cycle is
needed before comparable standardized recursive residuals are typically
available. For monthly data with `frequency = 12`, the residuals for roughly
the first year may therefore be `NA`.

The `NA` values in standardized recursive residuals should not be interpreted
as observations being omitted from likelihood evaluation. The log-likelihood is
computed by KFAS through the Kalman filtering likelihood with diffuse
initialization handled according to the selected likelihood type. In contrast,
standardized recursive residuals are diagnostic quantities and may be
unavailable during the diffuse phase. Thus, likelihood-based fitting and
residual diagnostics use related but not identical quantities.

This distinction is important for residual diagnostics. ACF plots and
Ljung-Box diagnostics should be interpreted for the period where finite
standardized recursive residuals are available. Preserving the time index with
`get_tempssm_residuals()` helps identify which part of the original time series
is included in the residual diagnostic assessment.

In this example, the ACF plot does not show a clear sequence of successive
significant lags or a repeated seasonal pattern. Isolated spikes can occur when
many lags are inspected, so they should be interpreted together with the
Ljung-Box test and the residual time-series plot rather than used as the sole
basis for changing the model.

Use `plot_tempssm_residuals(resid)` to visualize the extracted residual series,
ACF plot, and residual frequency distribution. Use
`diagnose_residual_ts(resid)` to obtain a compact table with the Ljung-Box
diagnostic and residual kurtosis. The default Ljung-Box lag follows the
seasonal frequency of the input data; for monthly data, this uses lag 12.

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag <- diagnose_residual_ts(resid)
print(diag)
```

The `lb_stat`, `lb_lag`, and `lb_pvalue` columns correspond to the Ljung-Box
test statistic, the lag used in the test, and the P-value. For monthly time
series, the default lag is 12. In this example, the test does not indicate
significant residual autocorrelation up to lag 12.

If residual autocorrelation at longer lags is of concern, the lag can be
specified manually. For example, the following code repeats the Ljung-Box test
up to lags 24 and 36.

```{r message = FALSE, warning = FALSE, echo = TRUE}
diag_lag24 <- diagnose_residual_ts(resid, lb_lag = 24)
diag_lag36 <- diagnose_residual_ts(resid, lb_lag = 36)

diag_lags <- rbind(diag, diag_lag24, diag_lag36)
diag_lags$check <- c("lag 12", "lag 24", "lag 36")
diag_lags[, c("check", "lb_stat", "lb_lag", "lb_pvalue", "kurtosis")]
```

In this example, extending the Ljung-Box diagnostic to lags 24 and 36 still
does not indicate significant residual autocorrelation. Interpret these tests
together with the residual ACF plot. Isolated ACF spikes can occur by chance,
whereas repeated or periodic spikes may suggest remaining temporal structure.

### Examine the Autoregressive Order

This manual uses AR(2) as the baseline specification because the simulated SST
series was generated under an AR(2) setting informed by Baba et al. (2024). The
autoregressive component absorbs short-term serial dependence not represented
by the level, drift, and seasonal components.

The choice of AR order should still be checked against diagnostics. As a
sensitivity check, we also fit an AR(1) model and repeat the same residual
diagnostics. In applications without prior information about the AR structure,
AR(1) is a reasonable starting point, and higher-order alternatives such as
AR(2) can be examined when residual autocorrelation remains.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Optional sensitivity check with a first-order autoregressive component
res_ar1 <- tempssm(sst_sim, ar_order = 1)
summary(res_ar1)
resid_ar1 <- get_tempssm_residuals(res_ar1)
plot_tempssm_residuals(resid_ar1)
diag_lag12_ar1 <- diagnose_residual_ts(resid_ar1, lb_lag = 12)
diag_lag24_ar1 <- diagnose_residual_ts(resid_ar1, lb_lag = 24)
diag_lag36_ar1 <- diagnose_residual_ts(resid_ar1, lb_lag = 36)
diag_lags_ar1 <- rbind(diag_lag12_ar1, diag_lag24_ar1, diag_lag36_ar1)
diag_lags_ar1$check <- c("lag 12", "lag 24", "lag 36")
diag_lags_ar1[, c("check", "lb_stat", "lb_lag", "lb_pvalue", "kurtosis")]
```

In this example, the AR(1) diagnostics are broadly similar to those from the
AR(2) baseline model. We therefore proceed with AR(2), reflecting the
simulation setting, while noting that the main component patterns are not
highly sensitive to this AR-order choice in this example.

### Extract Estimated Components

The fitted `tempssm` object contains smoothed state estimates returned by
`KFAS`. The matrix `res_base$kfs$alphahat` stores the level, drift, seasonal
states, and autoregressive state.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimates
alpha_hat <- res_base$kfs$alphahat
head(alpha_hat)
```

For routine use, helper functions extract individual components as `ts`
objects with the original time index. The level component represents the
estimated long-term temperature level, while the drift component represents
its rate of change per year.

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Smoothing estimate of level component
level_ts <- get_level_ts(res_base)

# Smoothing estimate of drift component
drift_ts <- get_drift_ts(res_base)

# Average drift rate per year across the full period
mean_drift_year <- mean(drift_ts) 
print(mean_drift_year)
```
 
The average annual rate of SST change was estimated to be
`r sprintf("%.4f", mean_drift_year)` °C over the full period. Because the
estimated drift changes sign over time, this value should be interpreted as a
model-based summary of the whole study period rather than as a constant
warming rate.

### Make Short-Term Predictions

A fitted `tempssm` object can also be passed to `predict()` to obtain
short-term predictions. By default, `predict(res)` returns a one-step-ahead
prediction beyond the end of the observed series. This is useful for visual
checks of how the fitted model extrapolates the estimated level, seasonal, and
autoregressive components.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_1 <- predict(res_base)
pred_1
```

Predictions for multiple future time points can be requested by setting the
`n.ahead` argument.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12 <- predict(res_base, n.ahead = 12)
pred_12
```

Prediction uncertainty can also be returned by setting the `interval`
argument. With `interval = "confidence"`, the returned lower and upper bounds
describe uncertainty in the predicted mean response. With
`interval = "prediction"`, the bounds describe uncertainty in a future
observation and therefore also include observation error; prediction intervals
are usually wider than confidence intervals. The confidence level is controlled
by the `level` argument and defaults to 0.95.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_12_pi <- predict(
  res_base,
  n.ahead = 12,
  interval = "prediction",
  level = 0.95
)

pred_12_pi
```

These predictions should be interpreted as model-based extrapolations rather
than definitive forecasts. Uncertainty generally increases as the prediction
horizon becomes longer, and long-horizon predictions can be sensitive to model
assumptions about trend, seasonality, and autoregressive dependence.


## Exercise II: Model With Exogenous Variables

We extend the state-space modeling framework to examine the effect of
exogenous factors on temperature variation. Specifically, we use simulated
marine-environment variables related to the Kuroshio path as exogenous
variables for the simulated SST series introduced in Exercise I.

### Load Simulated Marine-Environment Variables

Two simulated exogenous-variable datasets are included in the package.

- **`kuroshio_a_sim`**: monthly binary indicator of the Kuroshio A-type path  
- **`distance_sim`**: monthly offshore-distance index of the Kuroshio path  
- **Period**: January 1998 to February 2023

Both datasets are provided as univariate `ts` objects with `frequency = 12`.
They are simulated monthly datasets based on the Kuroshio-related variables
used in Baba et al. (2024). The original data used in that study had finer
time resolution for some variables, but the simulated datasets in this package
are represented as monthly time series for use with `tempssm` examples.

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(kuroshio_a_sim)
data(distance_sim)

head(kuroshio_a_sim)
head(distance_sim)

summary(kuroshio_a_sim)
summary(distance_sim)
```

### Check Time-Series Alignment

For state-space modeling with exogenous variables, the response and exogenous
time series must share a common and aligned time index. In this example,
`sst_sim`, `kuroshio_a_sim`, and `distance_sim` already have the same monthly
time span and frequency.

```{r message = FALSE, warning = FALSE, echo = TRUE}
series_info <- tibble::tibble(
  series = c("sst_sim", "kuroshio_a_sim", "distance_sim"),
  start = c(
    paste(start(sst_sim), collapse = "-"),
    paste(start(kuroshio_a_sim), collapse = "-"),
    paste(start(distance_sim), collapse = "-")
  ),
  end = c(
    paste(end(sst_sim), collapse = "-"),
    paste(end(kuroshio_a_sim), collapse = "-"),
    paste(end(distance_sim), collapse = "-")
  ),
  frequency = c(
    frequency(sst_sim),
    frequency(kuroshio_a_sim),
    frequency(distance_sim)
  ),
  missing = c(
    sum(is.na(sst_sim)),
    sum(is.na(kuroshio_a_sim)),
    sum(is.na(distance_sim))
  )
)

series_info
```

If user-supplied time series have different start or end points,
`trim_ts_overlap()` can be used to restrict the response and exogenous series
to their shared overlapping period before fitting the model.

### Plot the Response and Exogenous Variables

We visualize the simulated SST series together with the two simulated Kuroshio
variables to inspect the overall structure of the dataset.

```{r message = FALSE, warning = FALSE, echo = TRUE}
plt_sst_exo <- forecast::autoplot(sst_sim) +
  labs(y = expression(Temp.~(degree*C)), x = "Time (year)") +
  ggtitle("Simulated monthly SST") +
  theme_classic()

plt_kuroshio_a <- forecast::autoplot(kuroshio_a_sim) +
  labs(x = "Time (year)", y = "A-type indicator") +
  ggtitle("Simulated Kuroshio A-type path") +
  scale_y_continuous(breaks = c(0, 1)) +
  theme_classic()

plt_distance <- forecast::autoplot(distance_sim) +
  labs(x = "Time (year)", y = "Offshore distance") +
  ggtitle("Simulated Kuroshio offshore distance") +
  theme_classic()

plt_sst_exo + plt_kuroshio_a + plt_distance +
  patchwork::plot_layout(ncol = 1)
```

The simulated exogenous variables used here contain no missing values. When
using your own exogenous data, resolve missing values before model fitting by,
for example, imputing them, excluding missing periods, or adjusting the
analysis period so that no `NA` values remain in the exogenous series.
`tempssm()` does not allow missing exogenous values. In contrast, missing
values in the dependent temperature series are allowed and are treated as
unobserved responses during Kalman filtering and smoothing.

### Reuse the Baseline Model

Exercise II uses the same response series as Exercise I, so we reuse the
previously fitted AR(2) baseline model as the reference model without exogenous
variables.

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_without <- res_base
summary(res_without)
```

This keeps the comparison focused on the additional information introduced by
the simulated Kuroshio variables.

### Combine Exogenous Variables

The `exo_data` argument of `tempssm()` can accept either a univariate `ts`
object or a multivariate `ts` object. Because this exercise uses two exogenous
variables, we combine `kuroshio_a_sim` and `distance_sim` into a single
multivariate `ts` object before fitting the model.

```{r message = FALSE, warning = FALSE, echo = TRUE}
exo_kuroshio <- cbind(kuroshio_a_sim, distance_sim)
colnames(exo_kuroshio) <- c("kuroshio_a", "distance")

head(exo_kuroshio)
start(exo_kuroshio)
end(exo_kuroshio)
frequency(exo_kuroshio)
colnames(exo_kuroshio)
sum(is.na(exo_kuroshio))
```

The resulting object has class `mts`, the multivariate form of base R's `ts`
class. Each column is treated as a separate exogenous variable, and the column
names are used as labels in summaries and coefficient tables.

### Fit a Model With Exogenous Variables

We next fit a model with the two simulated Kuroshio variables supplied through
`exo_data`.

```{r message = FALSE, warning = FALSE, echo = TRUE}
res_with <- tempssm(
  temp_data = sst_sim,
  exo_data = exo_kuroshio,
  ar_order = 2
)
summary(res_with)

resid_with <- get_tempssm_residuals(res_with)
plot_tempssm_residuals(resid_with)
diag_res_with <- diagnose_residual_ts(resid_with)
print(diag_res_with)
```

```{r echo = FALSE}
exo_coef <- get_exo_coef(res_with)
kuroshio_a_coef <- exo_coef$Coefficient[exo_coef$Variable == "kuroshio_a"]
kuroshio_a_lwr <- exo_coef$lwr[exo_coef$Variable == "kuroshio_a"]
kuroshio_a_upr <- exo_coef$upr[exo_coef$Variable == "kuroshio_a"]
distance_coef <- exo_coef$Coefficient[exo_coef$Variable == "distance"]
distance_lwr <- exo_coef$lwr[exo_coef$Variable == "distance"]
distance_upr <- exo_coef$upr[exo_coef$Variable == "distance"]
```

The residual time-series plot shows that standardized recursive residuals are
`NA` until 2004, corresponding to the period before the diffuse components
from diffuse initialization have been resolved. The residual diagnostics for
this model also do not indicate significant residual autocorrelation up to lag
12. Thus, the AR(2) specification remains acceptable for both models
considered here.

We then examine the estimated coefficients for the exogenous variables.

```{r message = FALSE, warning = FALSE, echo = TRUE}
exo_coef
```

The coefficient table summarizes the estimated association between SST and
each simulated Kuroshio variable after accounting for the latent components.
In this fitted model, the estimated coefficient for `kuroshio_a` is
`r sprintf("%.2f", kuroshio_a_coef)` °C, with a 95% confidence interval from
`r sprintf("%.2f", kuroshio_a_lwr)` to
`r sprintf("%.2f", kuroshio_a_upr)`. Because `kuroshio_a` is coded as a binary
indicator, this coefficient can be read as the estimated SST difference
between A-type and non-A-type conditions, conditional on the other model
components and on `distance`.

The estimated coefficient for `distance` is
`r sprintf("%.3f", distance_coef)` °C per one-unit increase in the distance
index, with a 95% confidence interval from
`r sprintf("%.3f", distance_lwr)` to `r sprintf("%.3f", distance_upr)`. This
means that larger values of the simulated distance index are associated with
slightly lower SST, conditional on the latent components and `kuroshio_a`.

Because these are simulated example datasets, these coefficient
interpretations should be treated as an illustration of how to fit and inspect
a model with multiple exogenous variables, rather than as new scientific
evidence about the Kuroshio path.

For models with exogenous variables, future exogenous values are needed when
calling `predict()`. They can be supplied with `new_exo_data`, using the same
variables and column order as `exo_data`. For a simple one-step-ahead visual
check, `exo_strategy = "last"` carries the final observed row forward as a
persistence assumption.

```{r message = FALSE, warning = FALSE, echo = TRUE}
pred_with_last_exo <- predict(res_with, exo_strategy = "last")
pred_with_last_exo
```

### Compare Estimated Components

We compare the estimated long-term trend and drift between the two models by
overlaying their smoothed point estimates.

```{r message = FALSE, warning = FALSE, echo = TRUE}
level_without <- get_level_ts(res_without)
level_with <- get_level_ts(res_with)
drift_without <- get_drift_ts(res_without)
drift_with <- get_drift_ts(res_with)

component_compare <- rbind(
  data.frame(
    time = as.numeric(time(level_without)),
    component = "Level component (°C)",
    model = "Without exogenous variables",
    estimate = as.numeric(level_without)
  ),
  data.frame(
    time = as.numeric(time(level_with)),
    component = "Level component (°C)",
    model = "With Kuroshio variables",
    estimate = as.numeric(level_with)
  ),
  data.frame(
    time = as.numeric(time(drift_without)),
    component = "Drift component (°C/year)",
    model = "Without exogenous variables",
    estimate = as.numeric(drift_without)
  ),
  data.frame(
    time = as.numeric(time(drift_with)),
    component = "Drift component (°C/year)",
    model = "With Kuroshio variables",
    estimate = as.numeric(drift_with)
  )
)

component_compare$component <- factor(
  component_compare$component,
  levels = c("Level component (°C)", "Drift component (°C/year)")
)

plt_component_compare <- ggplot(
  component_compare,
  aes(x = time, y = estimate, color = model)
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ component, ncol = 1, scales = "free_y") +
  labs(x = "Time (year)", y = "Estimate", color = "Model") +
  theme_classic()

plot(plt_component_compare)

mean_drift_year_without <- mean(drift_without, na.rm = TRUE)
mean_drift_year_with <- mean(drift_with, na.rm = TRUE)

mean_drift_year_without
mean_drift_year_with
```

The level components show a similar broad pattern in both models: SST declines
from the beginning of the record to around 2008 and then increases toward the
end of the period. However, the model with the simulated Kuroshio variables
estimates a higher level during the early period and a slightly lower level
toward the end, compared with the model without exogenous variables.

The difference between the two models is clearer in the drift component. The
model without exogenous variables shows more local fluctuation in the rate of
change, whereas the model with the simulated Kuroshio variables gives a
smoother transition from negative to positive drift. This suggests that some
short- to medium-term variation represented in the baseline model's drift
component is instead captured by the simulated exogenous variables.

The estimated annual rate of SST change over the observation period is
`r sprintf("%.4f", mean_drift_year_without)` in the model without exogenous
variables and `r sprintf("%.4f", mean_drift_year_with)` in the model with
exogenous variables.

This simulated example illustrates that estimated long-term components can
change when external drivers are included.

The next section evaluates whether this difference is also reflected in
out-of-sample predictive performance under a conditional forecasting setting.

### Evaluate Conditional Predictive Performance With tsCV

Time-series cross-validation (tsCV) evaluates out-of-sample prediction errors
while respecting temporal order. The model is repeatedly fitted to training
windows, and predictions are evaluated on subsequent observations, avoiding
information leakage from the future to the past.

Here, by comparing cross-validation metrics for models with and without the
simulated Kuroshio variables, we assess whether the additional exogenous
information is reflected in out-of-sample predictive performance.

For the model with the simulated Kuroshio variables, the test-period values of
those variables are supplied to the forecast step as exogenous variables.
Therefore, this tsCV setting evaluates conditional forecasts, or hindcasts,
under the assumption that the Kuroshio-variable values during each test period
are known. This is appropriate when the aim is to assess the explanatory or
reconstructive value of the exogenous variables.

For real-time future forecasting, however, future exogenous-variable values are
usually unknown. In that setting, users must either provide future scenarios,
forecast the exogenous variables with separate models, or use a simplified
assumption such as carrying the last observed value forward. The results below
should therefore be interpreted as evidence for improved conditional
predictive performance, not as a direct guarantee of operational forecast
improvement.


```{r message = FALSE, warning = FALSE, echo = TRUE}

## Generate a list of training and test datasets with their indices
# Procedure for constructing year-based time-series cross-validation folds:
# 
# First training data: January 1998-December 2010;
# First test data: January 2011-December 2011
# 
# Second training data: January 1999-December 2011;
# Second test data: January 2012-December 2012
# ...
# 
# Twelfth training data: January 2009-December 2021;
# Twelfth test data: January 2022-December 2022
#
# These folds are automatically generated by the ts_train_test_split() function.
# The final two observations in 2023 are not used because allow_partial = FALSE
# keeps every test set as a complete 12-month period.

# Generate training and test dataset for model without exogenous variables
folds_without <- ts_train_test_split(
  temp_data = sst_sim,
  exo_data = NULL,
  initial = 156, # 156 monthly observations from Jan 1998 to Dec 2010
  horizon = 12, # forecast 12 monthly observations
  step = 12, # move the fixed-width window in one-year steps
  fixed_window = TRUE,
  allow_partial = FALSE
  )

# Generate training and test dataset for model with Kuroshio variables
folds_with <- ts_train_test_split(
  temp_data = sst_sim,
  exo_data = exo_kuroshio,
  initial = 156, # 156 monthly observations from Jan 1998 to Dec 2010
  horizon = 12, # forecast 12 monthly observations
  step = 12, # move the fixed-width window in one-year steps
  fixed_window = TRUE,
  allow_partial = FALSE
  )

# Check the first fold
start(folds_without[[1]]$train_ts)
end(folds_without[[1]]$train_ts)
start(folds_without[[1]]$test_ts)
end(folds_without[[1]]$test_ts)


# *****************************************
# Executing time-series cross validation

#--------------------------------------------------
# Model without exogenous variables

# ts_cv_run() evaluates folds sequentially by default.
 cv_without_results <- ts_cv_run(
   folds_without,
   ar_order = 2,
   use_season = TRUE
 )

# Optional parallel execution:
# Calling library(future) or library(future.apply) is not required here,
# but these suggested packages must be installed.
#cv_without_results <- ts_cv_run(
#  folds_without,
#  ar_order = 2,
#  use_season = TRUE,
#  parallel = TRUE,
#  workers = 2
#)

# Computing assessment indexes
metrics_without <- lapply(cv_without_results, compute_cv_metrics)

# tidy summary
cv_without_tbl <- ts_cv_collect(cv_without_results, metrics_without) %>%
  mutate(Model="Without")


#--------------------------------------------------
# Model with simulated Kuroshio exogenous variables

 cv_with_results <- ts_cv_run(
   folds_with,
   ar_order = 2,
   use_season = TRUE
 )

# Optional parallel execution:
# Calling library(future) or library(future.apply) is not required here,
# but these suggested packages must be installed.
#cv_with_results <- ts_cv_run(
#  folds_with,
#  ar_order = 2,
#  use_season = TRUE,
#  parallel = TRUE,
#  workers = 2
#)

# Computing assessment indexes
metrics_with <- lapply(cv_with_results, compute_cv_metrics)

# tidy summary
cv_with_tbl <- ts_cv_collect(cv_with_results, metrics_with) %>%
  mutate(Model="With")


cv_tbl <- bind_rows(cv_without_tbl, cv_with_tbl)

cv_comparison <- compare_ts_cv(
  list(
    Without = cv_without_tbl,
    With = cv_with_tbl
  )
)

cv_comparison %>% knitr::kable()


plt_MAE <- ggplot(data=cv_tbl,
                  aes(x=Model,y=MAE)) +
  geom_boxplot()

plt_MASE_naive <- ggplot(data=cv_tbl,
                   aes(x=Model,y=MASE_naive)) +
  geom_boxplot()

plt_MASE_seasonal <- ggplot(data=cv_tbl,
                   aes(x=Model,y=MASE_seasonal)) +
  geom_boxplot()


plt_tsCV <- plt_MAE + plt_MASE_naive + plt_MASE_seasonal +
  patchwork::plot_layout(nrow = 1)

plot(plt_tsCV)

```

The three accuracy metrics summarize prediction errors from complementary
viewpoints. `MAE` is the mean absolute error and is expressed in degrees
Celsius. `MASE` is a unit-free scaled error measure relative to a simple
benchmark computed from the training data. Smaller values indicate better
predictive performance for all three metrics.

In this tutorial, `MASE_naive` uses a non-seasonal naive benchmark based on
one-step changes in the training series. `MASE_seasonal` uses a seasonal naive
benchmark based on changes separated by the seasonal frequency of the training
series; for monthly data, this corresponds to changes over 12 months. Values
below 1 indicate that the model performs better than the corresponding naive
benchmark on average.

```{r echo = FALSE}
monthly_mean_sst <- tapply(
  as.numeric(sst_sim),
  cycle(sst_sim),
  mean,
  na.rm = TRUE
)
seasonal_amplitude <- diff(range(monthly_mean_sst, na.rm = TRUE))
mae_without <- cv_comparison$mean_MAE[cv_comparison$model == "Without"]
mae_with <- cv_comparison$mean_MAE[cv_comparison$model == "With"]
mae_relative_to_season <- 100 * mae_with / seasonal_amplitude
mae_improvement <- 100 * (mae_without - mae_with) / mae_without
n_folds <- unique(cv_comparison$n_folds)
n_folds <- if (length(n_folds) == 1L) n_folds else min(n_folds)
```

The comparison table produced by `compare_ts_cv()` summarizes the number of
folds, convergence rates, and mean prediction-error metrics for each model.
The boxplots show how fold-level errors vary across validation periods. In
this tutorial, tsCV uses `r n_folds` annual test periods, providing a practical
basis for comparing the two model specifications within this example dataset.
For formal model assessment, fold design should be chosen according to the
time-series length, prediction horizon, and intended forecasting task.

The mean MAE of the model with exogenous variables is approximately
`r sprintf("%.2f", mae_with)` °C. Relative to the seasonal amplitude based on
monthly mean SST (about `r sprintf("%.1f", seasonal_amplitude)` °C), this
corresponds to roughly `r sprintf("%.1f", mae_relative_to_season)`%. Thus, the
prediction error is small compared with the dominant seasonal scale of
variation in this data set, and the model can be considered practically
accurate depending on the intended purpose.

The model with the exogenous variables reduces the mean MAE by approximately
`r sprintf("%.1f", mae_improvement)`% compared with the model without
exogenous variables. Similar improvements are also seen in the MASE metrics.
This interpretation should be kept conditional on the tsCV setting used here,
in which the test-period values of the simulated Kuroshio variables are
supplied as exogenous variables.

Together, the coefficient estimates, component comparison, and tsCV results
illustrate how exogenous-variable models can be assessed. Because the datasets
are simulated, the emphasis is on the workflow rather than on drawing new
scientific conclusions about Kuroshio variability. The same workflow can be
extended to other environmental, climatic, or ecological covariates, while
keeping model diagnostics, predictive performance, and forecasting assumptions
explicit.

# Using Your Own Data

## CSV Format
For monthly temperature data stored in a CSV file, prepare columns named
`Year`, `Month`, and `Temp`.

- Year
- Month
- Temp

Example:
```text
Year,Month,Temp
2010,8,13.6
2010,9,6.8
2010,10,NA
2010,11,-1.4
...
```

* Use NA for missing temperature values, and always keep the corresponding Year
  and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.


## Convert CSV to a `ts` Object

An example CSV file included in this package is available in inst/extdata. The
example dataset contains monthly air temperature observations from Mt. Akadake,
Hokkaido, Japan (1,840 m). The original data are available from the Monitoring
Sites 1000 Project of the Ministry of the Environment of Japan (KOZ01.zip,
https://www.biodic.go.jp/moni1000/findings/data/index.html).

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")
akadake_temp <- tempssm::read_monthly_temp_ts(path)
head(akadake_temp)
```

The `read_monthly_temp_ts()` function reads this type of CSV file and converts
it into an R `ts` object for use with `tempssm()`. If finer control is needed,
you can also construct a `ts` object manually; see the `stats::ts`
documentation for details:
<https://search.r-project.org/R/refmans/stats/html/ts.html>.

If your data are already stored as an R `ts` object, you can pass them
directly to `tempssm()`.



# References

The statistical modeling framework implemented in `tempssm` is based on the
methodology described in Baba et al. (2024), and the accompanying supplementary
materials and code repository served as the initial source for development of
this package.

Baba, S., Ishii, H., and Yoshiyama, T. (2024).
Estimating sea temperature trends using a linear Gaussian state-space model
in Jogashima, Kanagawa, Japan.
*Bulletin of the Japanese Society of Fisheries Oceanography*,
88(3), 190–199.
(In Japanese with an English abstract.)
https://doi.org/10.34423/jsfo.88.3_190

Baba, S. (2024).
Supplementary code and test data for estimating sea temperature trends using a
linear Gaussian state-space model.
GitHub repository:
https://github.com/logics-of-blue/sea-temperature-trend-jogashima

Helske, J. (2017).
KFAS: Exponential Family State Space Models in R.
*Journal of Statistical Software*, 78(10), 1–39.
https://doi.org/10.18637/jss.v078.i10

# Appendix: Utility Functions

The following utility functions are provided to support data preparation and
exploratory analysis.


## 1. `read_monthly_temp_ts()`

The `read_monthly_temp_ts()` function converts monthly temperature data stored
in a CSV file into an R `ts` object. By enforcing a simple and consistent data
format, it is intended to make externally prepared time-series data easier to
import into tempssm.

### Example

For monthly temperature data, prepare a CSV file with a header row. By default,
the column names must be `Year`, `Month`, and `Temp`, where `Temp` represents
the observed temperature value.

```text
Year,Month,Temp
2001,1,10.4
2001,2,8.2
2001,3,NA
2001,4,13.6
2001,5,16.1
...
```

* Use NA for missing temperature values, and always keep the corresponding Year
  and Month entries.
* The CSV file must be comma-separated and UTF-8 encoded.

The following example uses the sample CSV file included in the package. It
contains monthly air temperature observations from Mt. Akadake, Hokkaido,
Japan.

```{r message = FALSE, warning = FALSE, echo = TRUE}
path <- system.file("extdata", "example_monthly_temp.csv", package = "tempssm")

# Read the CSV file and convert it to a monthly ts object
temp_ts <- read_monthly_temp_ts(path)
head(temp_ts)
```

## 2. `convert_monthly_df_to_ts()`

The `convert_monthly_df_to_ts()` function converts a data frame containing
monthly temperature data into an R `ts` object. It is designed to support
workflows in which temperature data are first imported or prepared as a data
frame and then used for time-series analysis.

The input data frame is expected to contain at least two columns: a date column
(`Date`) and a temperature column (`Temp`). For the date column (`Date`), assign
an arbitrary day of the month, such as the first day, so that the dates
correspond to monthly aggregated temperature values.

### Example

```{r message = FALSE, warning = FALSE}
# Create a data frame of monthly temperature data
df <- data.frame(
  Date = as.Date(c(
    "2001-01-01",
    "2001-02-01",
    "2001-03-01",
    "2001-04-01",
    "2001-05-01")
    ),
  Temp = c(10.4, 8.2, NA, 13.6, 16.1)
  )

# Convert to a monthly ts object
temp_ts <- convert_monthly_df_to_ts(df)
head(temp_ts)
```

## 3. `get_jma_sst_ts()`

The `get_jma_sst_ts()` function downloads daily mean sea surface temperature
(SST) data for Japanese coastal waters provided by the Japan Meteorological
Agency (JMA). It aggregates the daily values into monthly means and returns the
resulting time series as an object of class `ts`.

### Example

The following example demonstrates how to download SST data for the southern
coastal waters of Ibaraki Prefecture, Japan.

The argument `sea_area_id` specifies the numeric identifier of a sea area
defined by JMA. For example, `138` corresponds to the coastal waters off
southern Ibaraki. A list of available sea area IDs and their corresponding
regions is provided by JMA at:

<https://www.data.jma.go.jp/kaiyou/data/db/kaikyo/series/engan/eg_areano.html>

```{r message = FALSE, warning = FALSE, echo = TRUE}
sst_138_ts <- get_jma_sst_ts(sea_area_id = 138)
head(sst_138_ts)
```

## 4. `compute_temp_anomaly()`

The `compute_temp_anomaly()` function computes anomalies from a time series
provided as an R `ts` object. Anomalies are calculated by subtracting the
corresponding long-term mean for each periodic unit, such as month, from each
observation. This transformation is useful for examining interannual
variability after removing periodic effects.

The reference period used to compute anomalies can be specified via a function
argument. By default, the climatology is computed over the entire available
time series (see `?compute_temp_anomaly`).

This transformation is useful for exploratory analysis and modeling
applications that focus on departures from typical seasonal conditions.

### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
# Generate temperature anomalies
data(sst_niigata)
sst_niigata_anomaly <- compute_temp_anomaly(sst_niigata)
plt_sst_niigata_anomaly <- forecast::autoplot(sst_niigata_anomaly) +
  labs(y = expression(Temp.~(degree*C)), 
       x = "Time (year)")
plot(plt_sst_niigata_anomaly) 
```

## 5. `compute_monthly_climatology()`

The `compute_monthly_climatology()` function computes the climatological mean
seasonal cycle from a `ts` object by averaging each seasonal value across
years. It is primarily intended for exploratory analysis and visualization of
seasonal structure in temperature time series.

### Example

```{r message = FALSE, warning = FALSE, echo = TRUE}
data(sst_niigata)
monthly_seasonal_cycle_sst_niigata <- compute_monthly_climatology(sst_niigata) 
summary(monthly_seasonal_cycle_sst_niigata)

plt_monthly_seasonal_cycle_sst_niigata <- ggplot(
  data = monthly_seasonal_cycle_sst_niigata,
  aes(x = Month, y = Temperature)
) +
  geom_point(size = 2) +
  geom_line(linetype = "dashed") +
  labs(
    title = "Monthly seasonal cycle of temperature",
    x = "Month",
    y = expression(Temp.~(degree*C))
  ) +
  scale_x_continuous(
    breaks = 1:12,
    labels = sprintf("%02d", 1:12)
  ) +
  theme_classic()


plot(plt_monthly_seasonal_cycle_sst_niigata)
```
